---
title: "Organising principles among vital rates: Reduced models (without time interval)"
subtitle: ""
author: "Melina Leite"
date: "`r format(Sys.time(), '%d de %B de %Y')`"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile,encoding=encoding, output_dir = "..") })
output:
  rmdformats::readthedown:
    highlight: kate
    self_contained: true
    thumbnails: false
    lightbox: true
    gallery: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = T, warning = F, fig.align = "center", message = F,
  error = F, cache = F, cache.lazy = FALSE)

library(knitr)
library(tidyverse); library(cowplot); library(patchwork); 
theme_set(theme_cowplot())
library(ggrepel); library(ggbeeswarm)
library(janitor)
library(pals) # paletas bem divergentes
library(DirichletReg)
library(htmlTable)
library(here)
library(sf)# map
library(rnaturalearth); library(rnaturalearthdata)# map
library(wesanderson) #better colour palette
```

# Data

Organising all tables of models results for all 21 forest plo in one object per vital rate, quadrat size scale.

See table S1 for forest plot names.

```{r run, echo=F}
#Pre-analysis 
# Gradient color para latitutde
pal <- wes_palette("Zissou1",100, type = "continuous")[100:1] # azul frio-temperado
#Figures colors/labels
cores <- c("#c00000", "#b666d2", "#70ad47", "#ffc000")

labvit <- as_labeller(c(grow = "Growth", mort="Mortality", 
                        rec="Recruitment"))
grplab = as_labeller(c(quadrat = "space", `quadrat:sp`= "space*species",
                        sp="species", Residual = "residual"))
grpvit <- as_labeller(c(grow = "Growth", mort="Mortality", 
                        rec="Recruitment",
                       quadrat = "space", `quadrat:sp`= "species*space",
                        sp="species", Residual = "residual"))
```


## Growth


Import results data to list  
```{r}
path = here("models_outputs/models_notime/all", "grow", "table")
files = list.files(path)

all = lapply(files, function(x) get(assign(x, load(paste0(path, "/", x)))))
tudo <- bind_rows(all)
```

Remove intercept to include again as a separate column
```{r}
intercepts <- tudo %>% filter(term == "intercept")
names(intercepts)[8] = "intercept"

grow <- tudo %>% filter(term != "intercept") %>%
  left_join(intercepts[,c(1:4,8)], by= c("data", "fplot", "time", "q_size")) %>%
  mutate(q_size = fct_relevel(q_size,"quad_5", "quad_10","quad_20", 
                              "quad_50", "quad_100")) %>%
  group_by(data, fplot, time, q_size) %>% 
  mutate(VPC = variance/sum(variance))  %>%
  as.data.frame()
```

Divide the sd of the terms by the mean growth (intercept) for each dataset.
```{r}
grow$stdev = grow$Estimate/grow$intercept
```

Number of growth models per florest plot and quadrat scale.Some forest plots have more than 1 model because we ran one model per census interval.
```{r}
table(grow$fplot,grow$q_size)/4
```

## Mortality

Import results data to list
```{r}
path = here("models_outputs/models_notime/all", "mort", "table")
files = list.files(path)

all = lapply(files, function(x) get(assign(x, load(paste0(path, "/", x)))))
tudo <- bind_rows(all)
```

Remove intercept to include again as a separate column
```{r}
intercepts <- tudo %>% filter(term == "intercept")
names(intercepts)[8] = "intercept"

mort <- tudo %>% filter(term != "intercept") %>%
  left_join(intercepts[,c(1:4,8)], by= c("data", "fplot", "time", "q_size")) %>%
  mutate(q_size = fct_relevel(q_size,"quad_5", "quad_10","quad_20", 
                              "quad_50", "quad_100")) %>%
  group_by(data, fplot, time, q_size) %>% 
  mutate(VPC = variance/sum(variance))  %>%
  as.data.frame()

mort$stdev <- mort$Estimate
```

Number of mortality models per florest plot and quadrat scale.Some forest plots have more than 1 modelo because we ran one model per census interval.
```{r}
table(mort$fplot,mort$q_size)/4
```

## Recruitment

Import results data to list   
```{r}
path = here("models_outputs/models_notime/all", "rec", "table")
files = list.files(path)

all = lapply(files, function(x) get(assign(x, load(paste0(path, "/", x)))))
tudo <- bind_rows(all)
```

Remove intercept to include again as a separate column
```{r}
intercepts <- tudo %>% filter(term == "intercept")
names(intercepts)[8] = "intercept"

rec <- tudo %>% filter(term != "intercept") %>%
  left_join(intercepts[,c(1:4,8)], by= c("data", "fplot", "time", "q_size")) %>%
  mutate(q_size = fct_relevel(q_size,"quad_5", "quad_10","quad_20", 
                              "quad_50", "quad_100")) %>%
  group_by(data, fplot, time, q_size) %>% 
  mutate(VPC = variance/sum(variance))  %>%
  as.data.frame()

rec$stdev <- rec$Estimate
```

Number of recruitment models per florest plot and quadrat scale. Some forest plots have more than 1 modelo because we ran one model per census interval.
```{r}
table(rec$fplot,rec$q_size)/4
```

Excluding WYW recruitment data because it has very low recruitment rates in all 3 intervals, 0.0001, 0.0004, 0.0007, respectivelly. 
```{r}
rec <- rec %>% filter(fplot!="wyw")
```


## Combining models' results

All vita rates
```{r}
vital <- bind_rows(list(grow = grow, mort = mort, rec  = rec), 
                   .id="vital") %>%
  mutate(q_size = fct_recode(q_size,`5`="quad_5",`10`="quad_10",`20`="quad_20",
                             `50`="quad_50", `100`="quad_100")) %>%
  mutate(q_size = fct_relevel(q_size,"5", "10", "20", "50", "100"),
         term = fct_relevel(term, "quadrat", "quadrat:sp", "sp", "Residual")) 
```

## Plot information

Richness by rarefaction at the 6ha plot size (sampling unit 20x20m).
```{r}
load(here("data", "rarefaction.curves.Rdata"))
rich.rare <- rare[rare$sites ==150,2:3]
colnames(rich.rare)[1] <- "richness.rarefaction"
vital <- vital %>% left_join(rich.rare, by="fplot")
```

## Averages

To get ONE value per forest-vital, first I calculate the mean of the posterior variances for the plots >1 interval. Then I use this value as the Variance and calculate the Variance Partition Coefficient (VPC).
```{r mvital}
mvital <- vital %>% group_by(vital,fplot, term, q_size) %>% 
                    summarise(variance = mean(variance),
                              stdev = mean(stdev),
                              richness.rarefaction = mean(richness.rarefaction),
                              ntrees = mean(ntrees)) %>%
  group_by(vital, fplot, q_size) %>%
  mutate(VPC = variance/sum(variance))
```

Mean VPC across all plots
```{r avital}
avital <- mvital %>% group_by(vital, term, q_size) %>% 
                    summarise(mean.stdev = mean(stdev),
                              mean.var = mean(variance),
                              mean.VPC = mean(VPC),
                              sd.VPC = sd(VPC)) 
```

TOTAL VPC across plots
```{r tvital}
tvital <- mvital %>% group_by(vital, term, q_size) %>% 
                    summarise(variance = sum(variance),
                              stdev = sqrt(variance)) %>%
  group_by(vital, q_size) %>%
  mutate(VPC = variance/sum(variance))
```


Saving objects for models results
```{r}
save(mvital, vital, file= here("models_outputs", "all_results_notime.Rdata"))
```


# Standard deviations

Stacked SD for all forest plots
```{r, fig.height=6, fig.width=12}
mvital %>%
  ggplot(aes(x=fct_reorder(fplot,richness.rarefaction,max), y=stdev, fill=term,col=term)) +
  geom_col(alpha=0.8) +
  scale_fill_manual(values= cores)+ scale_color_manual(values= cores)+
  facet_grid(q_size~vital, scales="free") +
  xlab("Number of species -->") +
  theme(legend.position="top",
        axis.text.x = element_text(angle=45, hjust=1))
```


## Standard deviations x species richness

5x5m quadrat scale
```{r}
overallSD <- mvital %>% group_by(vital, fplot, q_size, term, richness.rarefaction) %>%
  summarise(stdev = mean(stdev)) %>%
  group_by(vital, fplot, q_size,richness.rarefaction) %>%
  summarise(ALLstdev = sum(stdev))  %>%
  filter(q_size=="5")
```

Models SD with log(richness) per vital rate
```{r}
mg <- lm(ALLstdev ~ log(richness.rarefaction), 
         data=overallSD[overallSD$vital == "grow", ])
mm <- lm(ALLstdev ~ log(richness.rarefaction), 
         data=overallSD[overallSD$vital == "mort", ])
mr <- lm(ALLstdev ~ log(richness.rarefaction), 
         data=overallSD[overallSD$vital == "rec", ])
coefs <- bind_rows(list(grow=broom::tidy(mg), 
               mort=broom::tidy(mm),
               rec =broom::tidy(mr)), .id="vital")
vals = c(paste0("beta = ", round(coefs[2,3],2), "; p = ", round(coefs[2,6],3)),
paste0("beta = ", round(coefs[4,3],2), "; p = ", round(coefs[4,6],3)),
paste0("beta = ", round(coefs[6,3],2), "; p = ", round(coefs[6,6],3)))
anota <- data.frame(vital=c("grow", "mort", "rec"),
                    x=10, y=8,
                    lab = vals)
```

Overall standard deviations decreased with richness for recruitment.

```{r, fig.height=3, fig.width=8}
over <- overallSD %>%
ggplot(aes(x=richness.rarefaction, y=ALLstdev))+
  geom_point() +
  facet_grid(~vital, labeller = labvit) +
  scale_x_log10() +
  geom_smooth(method="lm", aes(linetype=vital)) +
  scale_linetype_manual(values=c("dashed", "dashed", "solid"))+
  theme(legend.position = "none",
        panel.background = element_rect(colour="lightgray"))+
  geom_text(data=anota, aes(x=x,y=y, label=lab), hjust=0) +
  xlab("Rarefied species richness (log)") +
  ylab("Standard deviation")
over
png(here("figures", "FIG_S5.3_SDxRicnhess.png"), height = 350, width=900, res=100)
over
dev.off()
```


Relationship between standard deviation for each term and species richness.

```{r}
mgrow <- lm(stdev ~ -1 + log(richness.rarefaction)*term, 
            data=mvital[mvital$vital == "grow",])
mmort <- lm(stdev ~ -1 + log(richness.rarefaction)*term, 
            data=mvital[mvital$vital == "mort" ,])
mrec <- lm(stdev ~ -1+log(richness.rarefaction)*term, 
           data=mvital[mvital$vital == "rec"  ,])
broom::glance(mgrow)
broom::glance(mmort)
broom::glance(mrec)
```

bonferroni correction -> 0.05/3 = 0.0166667

Taking coefficients for figure
```{r}
coefs <- bind_rows(list(grow=broom::tidy(mgrow), 
               mort=broom::tidy(mmort),
               rec =broom::tidy(mrec)), .id="vital") %>%
  filter(! term %in% c("termquadrat", "termquadrat:sp", "termsp", 
                       "termResidual")) %>%
  mutate(sig = ifelse(p.value<0.01666667, "yes", "no"),
         beta = paste0("beta = ", round(estimate,2), "; p = ", round(p.value,4)),
         xis = 150,
         y= 3)

coefs$term[coefs$term == "log(richness.rarefaction)"] <- "quadrat"
coefs$term[coefs$term == "log(richness.rarefaction):termsp"] <- "sp"
coefs$term[coefs$term == "log(richness.rarefaction):termquadrat:sp"] <- "quadrat:sp"
coefs$term[coefs$term == "log(richness.rarefaction):termResidual"] <- "Residual"
coefs <- coefs %>% mutate(term = fct_relevel(term, "quadrat","quadrat:sp", "sp", "Residual")) 

coefs$beta[coefs$term=="Residual" & coefs$vital !="grow"] <- ""
```

Figure
```{r, fig.height=8, fig.width=11}
mvital$line <- "dashed"
mvital$line[mvital$vital %in% c("grow", "rec") & mvital$term == "sp" ] <- "solid"
mvital$line[mvital$vital == "mort" & mvital$term == "quadrat:sp" ] <- "solid"
  
fig <- mvital %>% filter(q_size=="5") %>%
ggplot(aes(x=richness.rarefaction, y=stdev))+
  facet_grid(vital~term, labeller = grpvit) +
  geom_point()+
  scale_x_log10() +
  geom_smooth(method="lm", aes(linetype=line)) +
  scale_linetype_manual(values=c("dashed", "solid"))+
  geom_text(data=coefs, aes(x=xis,y=y, label=beta))+
  theme(legend.position = "none",
        panel.background = element_rect(colour="lightgray")) +
  xlab("Rarefied species richness (log)") + ylab("Standard deviation")
fig
png(here("figures", "FIG_S5.4_SD_regressions.png"), height = 700, width=900, res=100)
fig
dev.off()
```



# Variance Partition Coefficients VPC

Stacked VPC for all forest plots at the 5x5 m quadrat scale.
```{r, fig.height=8, fig.width=7}
vital %>% group_by(vital, fplot, q_size, term) %>%
  summarise(VPC = mean(VPC),
            richness.rarefaction = mean(richness.rarefaction),
            ntrees = mean(ntrees)) %>%
  filter(q_size == "5") %>%
  ggplot(aes(x=fct_reorder(fplot, richness.rarefaction,max ), 
             y=VPC, fill=term, col=term)) +
  geom_col(alpha=0.8) +
  scale_fill_manual(values= cores)+ 
  scale_color_manual(values= cores)+
  facet_wrap(~vital,ncol=1, labeller = labvit) +
  xlab("- Number of species +") +
  theme(axis.text.x = element_text(angle=45, hjust=1),
        legend.position="none")
```


```{r}
ah <- mvital %>% filter(q_size == "5") %>% 
  group_by(vital,term) %>%
  summarise(VPC_mean = as.character(round(mean(VPC), 2))) %>%
  mutate(x_val = as.numeric(VPC_mean), 
         y_val = 0.7:3.7)
ah[5,3] <-  c("0.10")
```


## Figure 1b: VPCs for all forest plots and vital rates 

At the 5x5m quadrat size scale.
```{r}
vpc.fig1 <- mvital %>% filter(q_size == "5") %>%
  ggplot(aes(x=VPC, y=rev(term), col=term)) + 
    facet_grid(~vital, labeller = labvit) +
  geom_jitter(height = 0.1, size=3, alpha=0.6) +
  scale_fill_manual(values=cores, labels=grplab) +
  scale_color_manual(values=cores, labels=grplab)  +
  scale_y_discrete(labels = c( "Residual","Species","Species x Space",  "Space")) +
  ylab("") +
  stat_summary(fun=mean, fun.min=mean, fun.max=mean, fatten=1, col="black",
                         geom="crossbar", width=0.4)+
  geom_vline(xintercept=c(0,0.2,0.4,0.6), col='gray', linetype="dashed", alpha=0.6)+
    geom_text(data=ah, aes(x=x_val, y=rev(y_val), label=VPC_mean),
              hjust=0.5,size=5,
              col="black")+
    theme(legend.position = "none",
          panel.background = element_rect(colour="lightgray"),
        axis.text.x = element_text(angle=45, hjust=1, size=16)) 
vpc.fig1
```

```{r}
avitsum <- avital %>% 
  filter(q_size == "5") %>%
  select(vital,term,mean.VPC, sd.VPC) %>% 
  pivot_wider(names_from = vital, values_from = c(mean.VPC, sd.VPC)) %>%
  select(term, mean.VPC_grow, sd.VPC_grow, mean.VPC_mort, sd.VPC_mort,
        mean.VPC_rec, sd.VPC_rec)

avitsum %>% mutate_if(is.numeric, round, digits=2) %>%
  htmlTable(header=c("" ,rep(c("Mean", "SD"), 3)),
            cgroup = c("", "Growth", "Mortality", "Recruitment"),
            n.cgroup =c(1,2,2,2),
            caption = "Average VPC values for models at the 5x5m quadrat grain size.")
```


## Figure 1c: Average VPCs

5x5 m quadrat size scale.
```{r}
tot.vpc.fig <- tvital %>% filter(q_size == "5") %>%
ggplot(aes(x=vital, fill=term, col=term, y=VPC)) + 
  geom_col(alpha=0.7,width = 0.6) +
  facet_grid(~vital, scales="free_x", labeller = labvit) +
   xlab("") + ylab("Mean VPC") +
    scale_fill_manual(values=cores, labels=grplab) +
  scale_color_manual(values=cores, labels=grplab) +
  theme(axis.text.x = element_blank(),
        axis.ticks = element_blank())
tot.vpc.fig
```


## Figure 3: Total VPC per vital rates across spatial grain sizes

```{r}
barvpc <- tvital %>% 
ggplot(aes(x=q_size, fill=term, col=term, y=VPC)) + 
  geom_col(alpha=0.7) +
  facet_grid(~vital, scales="free_x", labeller = labvit) +
   xlab("Spatial grain (quadrat size)") + ylab("VPC") +
    scale_fill_manual(values=cores, labels=grplab) +
  scale_color_manual(values=cores, labels=grplab) +
  theme(legend.position = "none") +
  labs(tag="(a)")
barvpc
png(here("figures", "FIG_3_VPC_notime_scales.png"), height = 550, width=1080, res=130)
barvpc
dev.off()
```

```{r}
tvital %>% select(vital, term, q_size, VPC) %>%
  pivot_wider(names_from = q_size, values_from = VPC) %>%
  mutate_if(is.numeric, round, digits=2) %>%
  htmlTable()
```


# Figure 1 complete

map
```{r}
###### xy locations ######
forest <- read.table(here("data/plot_sites_information.csv"), sep=";", header=T)
forest$long[forest$ID == "len"] <- forest$long[forest$ID == "len"] -2
fxy <- st_as_sf(forest[,-27], coords = c("long", "lat"), crs = 4326) #WSG84
fxy <- cbind(fxy, forest[,9:10])
not.yet <- c("cha")# remove not used plots
fxy$data.status = T
fxy$data.status[fxy$ID %in% not.yet ] <- F
fxy <- fxy %>% filter(data.status==T)
fxy$lat = st_coordinates(fxy)[,2]
fxy$long = st_coordinates(fxy)[,1]
fxy$site[fxy$site == "Smithsonian Environmental Research Center"] <- "SERC"
fxy$site[fxy$site == "Smithsonian Conservation Biology Institute"] <- "SCBI"

world <- ne_countries(scale = "medium", returnclass = "sf")
world <- world %>% filter(geounit != "Antarctica")
```

```{r, fig.height=20, fig.width=15}
set.seed(5)
A <- ggplot(data = world) +
  geom_sf(size=0.2) + 
  geom_sf(data=fxy, aes(fill=abs(lat)), size=4, shape=21, col="black") +
  geom_label_repel(data=fxy, aes(y=lat, x=long, label=site), size=5.5,
                   label.size = 0,
                   label.padding = 0.2) +
  scale_fill_gradientn(colours=pal) +
  scale_x_continuous(expand = expansion(mult=c(-0.12, -0.15))) +
  scale_y_continuous(expand = expansion(mult=-c(0.02, 0.15))) +
  xlab("") + ylab("") +
  theme_bw()+
  theme(legend.position= "none",
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        plot.margin =  margin(-0.5, 0.1, -1, -0.3, "cm")) +  labs(tag="(a)") +
  geom_hline(yintercept=c(23.5,0,-23.5), linetype="dashed", col="gray") +#+
  theme(plot.tag.position = c(0,1),
  plot.tag = element_text(size=20, vjust = 1.2, hjust = 0, face="bold"))

B <- vpc.fig1 + ggtitle("") + labs(tag = "(b)") +
  theme_cowplot()+
  theme(plot.margin =  margin(-0.5, 0.1, 0.2, 0, "cm"),
        legend.position= "none",
        #plot.tag.position = c(0,1),
       # plot.tag = element_text(vjust = 3, hjust = -1.2),
        text = element_text(size=20),
        axis.text = element_text(size=16))


C <- tot.vpc.fig + labs(tag = "(c)") + 
  theme_cowplot()+
  theme(plot.margin =  margin(0.3, 0, 0.2, 0.3, "cm"),
        legend.position= "none",
        #plot.tag.position = c(0,1),
        panel.spacing.x = unit(0.01, "cm"),
       # plot.tag = element_text(vjust = 3, hjust = 0),
        text = element_text(size=20),
        axis.ticks = element_blank(),
        axis.text.y = element_text(size=18),
        axis.text.x = element_blank())
bot <- plot_spacer() + B + C + plot_layout(widths = c(0.00005,1,0.6))
#ggsave("figs/FIG_1BC.jpeg", height = 5.5, width=15)


A + wrap_plots(bot)+
plot_layout(ncol=1,widths = c(0.1,1), heights = c(1,0.6))
ggsave(here("figures/figure_1.jpeg"), height = 12, width=15)
```


# DIRICHLET regression


Including latitude for color
```{r}
load(here("data", "plots_structure.Rdata"))
vital <- vital %>% left_join(plots.structure[,1:2], "fplot")
```


## Growth

```{r}
predirig <-  vital %>% 
  filter(vital == "grow", q_size == "5") %>%
  select(fplot, time, term, VPC, richness.rarefaction, lat) %>%
  ungroup() %>% select(-time) %>%
  group_by(fplot, term) %>% summarise_all(mean)

diridatag <- predirig %>%
  pivot_wider(names_from = term, values_from = VPC) %>% ungroup() %>%
  mutate(log.rich = log(richness.rarefaction),
         log.rich.o=log.rich) %>%
  mutate_at(vars(log.rich), scale)

vpcg <- DR_data(diridatag[,c("quadrat", "sp", "quadrat:sp",  "Residual")])
#plot(vpcg)
```

```{r}
mgrow <- DirichReg(vpcg~log.rich, data=diridatag, model="alternative", base=4)
summary(mgrow)
```



Predictions
```{r, fig.height=3}
newdata <- expand.grid(log.rich = seq(min(diridatag$log.rich),
                                      max(diridatag$log.rich), length.out=10))
pred <- predict(mgrow, newdata = newdata, se=T)
colnames(pred) <- colnames(vpcg)
newdata$log.rich.o <- newdata$log.rich*sd(diridatag$log.rich.o) +
  mean(diridatag$log.rich.o)
newdata$rich.o <- exp(newdata$log.rich.o)
newpredg <- cbind(newdata,pred) %>% pivot_longer(`quadrat`:`Residual`, names_to="term", values_to="pred")

newpredg %>% mutate(term = fct_relevel(term, "quadrat", "quadrat:sp", "sp",
                                     "Residual")) %>%
ggplot(aes(x=rich.o, y=pred, col=term)) +
  geom_line() + facet_grid(~term) +
  geom_point(data=predirig, aes(x=richness.rarefaction, y=VPC, col=term)) +
  scale_x_log10() +
  ggtitle("Mod log.rich")
```
Residuals
```{r, fig.height=3}
resid <- residuals(mgrow, type = "standardized")
resid <-  data.frame(resid[,1:4])
resid$log.rich <- diridatag$log.rich
resid <- resid %>% pivot_longer(1:4, names_to="term", values_to="resid")

ggplot(resid, aes(x=log.rich, y=resid)) + geom_point() +
facet_grid(~term)
```


## Mortality

```{r}
predirim <-  vital %>% 
  filter(vital == "mort", q_size == "5") %>%
  select(fplot, time, term, VPC,richness.rarefaction, lat) %>%
  ungroup() %>% select(-time) %>%
  group_by(fplot, term) %>% summarise_all(mean)
diridatam <- predirim %>%
  pivot_wider(names_from = term, values_from = VPC) %>% ungroup() %>%
  mutate(log.rich = log(richness.rarefaction),
         log.rich.o=log.rich) %>%
  mutate_at(vars(log.rich), scale)

vpcm <- DR_data(diridatam[,c("quadrat", "sp", "quadrat:sp",  "Residual")])
```

```{r, echo=T}
mmort <- DirichReg(vpcm~log.rich,data=diridatam, model="alternative", base=4)
summary(mmort)
```


```{r, fig.height=3}
newdata <- expand.grid(log.rich = seq(min(diridatam$log.rich),
                                      max(diridatam$log.rich), length.out=10))
pred <- predict(mmort, newdata = newdata, se=T)
confint(mmort)
colnames(pred) <- colnames(vpcm)
newdata$log.rich.o <- newdata$log.rich*sd(diridatam$log.rich.o) +
  mean(diridatam$log.rich.o)
newdata$rich.o <- exp(newdata$log.rich.o)
newpredm <- cbind(newdata,pred) %>% pivot_longer(`quadrat`:`Residual`, names_to="term", values_to="pred")

newpredm %>% mutate(term = fct_relevel(term, "quadrat", "quadrat:sp", "sp",
                                      "Residual")) %>%
  ggplot(aes(x=rich.o, y=pred, col=term)) +
  geom_line() + facet_grid(~term) +
  geom_point(data=predirim, aes(x=richness.rarefaction, y=VPC, col=term)) +
  scale_x_log10() +
  ggtitle("Mod log.rich")
```

Residuals
```{r, fig.height=3}
resid <- residuals(mmort, type = "standardized")
resid <-  data.frame(resid[,1:4])
resid$log.rich <- diridatam$log.rich
resid <- resid %>% pivot_longer(1:4, names_to="term", values_to="resid")

ggplot(resid, aes(x=log.rich, y=resid)) + geom_point() +
  facet_grid(~term)
```

## Recruitment

```{r}
predirir <-  vital %>% 
  filter(vital == "rec", q_size == "5") %>%
  select(fplot, time, term, VPC,richness.rarefaction, lat) %>%
  ungroup() %>% select(-time) %>%
  group_by(fplot, term) %>% summarise_all(mean)
diridatar<- predirir %>%
  pivot_wider(names_from = term, values_from = VPC) %>% ungroup() %>%
  mutate(log.rich = log(richness.rarefaction),
         log.rich.o=log.rich) %>%
  mutate_at(vars(log.rich), scale)

vpcr <- DR_data(diridatar[,c("quadrat", "sp", "quadrat:sp",  "Residual")])
```

```{r}
mrec <- DirichReg(vpcr~log.rich,data=diridatar, model="alternative", base=4)
summary(mrec)
```
Prediction
```{r, fig.height=3}
newdata <- expand.grid(log.rich = seq(min(diridatar$log.rich),
                                      max(diridatar$log.rich), length.out=10))
pred <- predict(mrec, newdata = newdata, se=T)
colnames(pred) <- colnames(vpcr)
newdata$log.rich.o <- newdata$log.rich*sd(diridatar$log.rich.o) +
  mean(diridatar$log.rich.o)
newdata$rich.o <- exp(newdata$log.rich.o)
newpredr <- cbind(newdata,pred) %>% pivot_longer(`quadrat`:`Residual`, names_to="term", values_to="pred")

newpredr %>% mutate(term = fct_relevel(term, "quadrat", "quadrat:sp", "sp",
                                      "Residual")) %>%
  ggplot(aes(x=rich.o, y=pred, col=term)) +
  geom_line() + facet_grid(~term) +
  geom_point(data=predirir, aes(x=richness.rarefaction, y=VPC, col=term)) +
  scale_x_log10() +
  ggtitle("Mod log.rich")
```

Residuals
```{r, fig.height=3}
resid <- residuals(mrec, type = "standardized")
resid <-  data.frame(resid[,1:4])
resid$log.rich <- diridatar$log.rich
resid <- resid %>% pivot_longer(1:4, names_to="term", values_to="resid")

ggplot(resid, aes(x=log.rich, y=resid)) + geom_point() +
  facet_grid(~term)
```

## Figure dirichlet 

Prediction intervals calculated in another code following the paper:

> Douma, J.C. & Weedon, J.T. (2019) Analysing continuous proportions in ecology and evolution: A practical introduction to beta and Dirichlet regression. Methods in Ecology and Evolution, 10, 1412–1430. https://doi.org/10.1111/2041-210X.13234.

```{r}
source(here("scripts", "prediction_intervals_dirichlet.R"), local=T)
```

```{r}
predis <- bind_rows(list(grow=newpredg, mort=newpredm, rec=newpredr),
                    .id="vital") %>% 
  mutate(term = fct_relevel(term, "quadrat", "quadrat:sp", "sp","Residual"))

pontos <- bind_rows(list(grow=predirig, mort= predirim, rec=predirir), .id="vital")%>% 
  mutate(term = fct_relevel(term, "quadrat", "quadrat:sp", "sp", "Residual"))
quants <- bind_rows(list(grow=quantg,mort=quantm, rec=quantr), .id="vital")%>% 
  mutate(term = fct_relevel(term, "quadrat", "quadrat:sp", "sp", "Residual"))
```

```{r}
# pvalues
res <- data.frame(vital = c("grow", "mort", "rec"),
                  P = NA,
                  term = "Residual")

test <- summary(mgrow)
vals <- as.data.frame(test$coef.mat)[1:6,]
vals$coef <- names(test$coefficients)[1:6]
valsg <- vals[ grep("log.rich", vals$coef),] %>% 
  mutate(term=c("quadrat", "sp", "quadrat:sp"))
test <- summary(mmort)
vals <- as.data.frame(test$coef.mat)[1:6,]
vals$coef <- names(test$coefficients)[1:6]
valsm <- vals[ grep("log.rich", vals$coef),] %>% 
  mutate(term=c("quadrat", "sp", "quadrat:sp"))
test <- summary(mrec)
vals <- as.data.frame(test$coef.mat)[1:6,]
vals$coef <- names(test$coefficients)[1:6]
valsr <- vals[ grep("log.rich", vals$coef),] %>% 
  mutate(term=c("quadrat", "sp", "quadrat:sp")) 

pvals <- bind_rows(list(grow=valsg,mort=valsm,rec=valsr), .id="vital") %>%
  dplyr::select(vital,`Pr(>|z|)`, term) %>% rename(P = `Pr(>|z|)`)
pvals <- rbind(pvals,res) %>% 
  mutate(term = fct_relevel(term, "quadrat", "quadrat:sp", "sp", "Residual"))

pvals$x <- 300
pvals$y <- 0.70
pvals$sig <- "ns"
pvals$sig[which(pvals$P < 0.016667)] <- paste0("p = ", round(pvals$P[which(pvals$P < 0.016667)],3))
pvals$sig[pvals$term=="Residual"] <- ""
pvals$sig[pvals$sig=="p = 0"] <- "p < 0.001"

```

```{r}
cfs <- bind_rows(list(grow=valsg,mort=valsm,rec=valsr), .id="vital") %>%
  select(vital, coef,term, Estimate, `Std. Error`,`Pr(>|z|)`)
ggplot(cfs, aes(x=Estimate, y=term, col=term)) +
  facet_grid(~vital) +
  geom_point() + 
  geom_linerange(aes(xmin=Estimate-`Std. Error`, xmax=Estimate+`Std. Error`))+
  geom_vline(xintercept = 0, linetype="dashed") +
  ggtitle("Log.Rich Estimates + Std.Error")
```

```{r}
library(wesanderson) #better colour palette
# Gradient color para latitutde
pal <- wes_palette("Zissou1",21, type = "continuous")[21:1] # azul frio-temperado
```


```{r}
fdiri_lat <-ggplot(predis, aes(x=rich.o, y=pred)) +
  geom_line(size=1) +
  facet_grid(vital~term, labeller=grpvit) +
  geom_smooth(data=quants,aes(x=rich.o, y=lower), se=F, size=0.1)+
  geom_smooth(data=quants,aes(x=rich.o, y=upper), se=F, size=0.1)  +
  geom_ribbon(data=quants,aes(x=rich.o, ymin=lower, ymax=upper,
                              y=mean), alpha=0.05, fill="blue",
              size=0)+
  geom_point(data=pontos, aes(x=richness.rarefaction, y=VPC, fill=abs(lat)),
             col="black", size=2.5, pch=21)+
  scale_fill_gradientn(colors=pal) +
  scale_y_continuous(minor_breaks = NULL) +
  scale_x_log10() +
  geom_text(data=pvals, aes(x=x,y=y, label=sig), size=3, hjust=1,vjust=0,
            col="black")+
  
  xlab("Species richness")+
  ylab("VPC") +
  theme_cowplot() +
  theme(panel.background = element_rect(colour="lightgray"),
        legend.position = "none")
fdiri_lat
png(here("figures", "FIG_4_dirichlet_regressions.png"), height = 600, width=800, res=100)
fdiri_lat
dev.off()
```
