---
title: "Results TIME models SUBSAMPLING plots 5ha"
author: "Melina Leite"
date: "`r  Sys.Date()`"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "..") })
output:
  rmdformats::readthedown:
    highlight: kate
    self_contained: true
    thumbnails: false
    lightbox: true
    gallery: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = T, warning = F, fig.align = "center", message = F,
  error = F, cache = F, cache.lazy = FALSE)

library(knitr)
library(tidyverse); library(cowplot); library(patchwork); 
theme_set(theme_cowplot())
library(ggrepel); library(ggbeeswarm)
library(janitor)
library(htmlTable)
library(here)
```

# Data

Organising all tables of models results for all6 forest plo in one object per vital rate, quadrat size scale.

See table S1 for forest plot names.

```{r run}
# Pre-analysis
#Figures colors/labels
cores <- c( quadrat =  "#c00000", 
           `sp:time`= "#203864",
            time = "#ed7d31",
           `quadrat:time` = "#843c0c", 
           `quadrat:sp`= "#b666d2", 
            sp = "#70ad47", 
            Residual="#ffc000")

labvit <- as_labeller(c(grow = "Growth", mort="Mortality", 
                        rec="Recruitment"))
grpTlab = as_labeller(c(quadrat = "Space", `sp:time`="Species x Time", 
                        `quadrat:time`="Space x Time", `quadrat:sp`= "Species x Space",
                        sp="Species", Residual = "Residual",
                        time= "Time"))
grpvitT <- as_labeller(c(grow = "Growth", mort="Mortality", 
                         rec="Recruitment",
                        quadrat = "Space", `sp:time`="Species x Time", 
                         `quadrat:time`="Space x Time", `quadrat:sp`= "Species x Space",
                         sp="Species", Residual = "Residual",
                        time= "Time"))
labplot = as_labeller(c(quadrat = "Space", `sp:time`="Species x Time", 
                `quadrat:time`="Space x Time", `quadrat:sp`= "Species x Space",
                        sp="Species", Residual = "Residual",
                        time= "Time",
                fus="Fushan", luq = "Luquillo", lam="Lambir", 
                pas = "Pasoh", bci="Barro Colorado"))
```


## Growth

Import results data to list  
```{r}
path = here("models_outputs/models_time/all", "grow", "table")
files = list.files(path)
 
all = lapply(files, function(x) get(assign(x, load(paste0(path, "/", x)))))

tudo <- bind_rows(all)
```

Remove intercept to include again as a separate column
```{r}
intercepts <- tudo %>% filter(term == "intercept")
names(intercepts)[8] = "intercept"

growT <- tudo %>% filter(term != "intercept") %>%
  left_join(intercepts[,c(1:4,8)], by= c("sub","data", "fplot", "q_size")) %>%
  group_by(sub, data, fplot, q_size) %>% 
  mutate(VPC = variance/sum(variance))  %>% ungroup() %>%
  mutate(term = fct_relevel(term, "quadrat", "sp:time","time",
                            "quadrat:time",
                            "quadrat:sp", "sp", "Residual"),
         q_size = fct_recode(q_size, `5`= "quad_5", `10`= "quad_10",
                             `20`="quad_20",
                            `50`="quad_50",  `100`="quad_100")) %>%
  mutate(q_size = fct_relevel(q_size, "5","10", "20","50", "100")) %>%
  as.data.frame()
```

Divide the sd of the terms by the mean growth (intercept) for each dataset.
```{r}
growT$stdev = growT$Estimate/growT$intercept
```


Number of growth models per florest plot and quadrat scale.
```{r}
table(growT$fplot, growT$q_size)/7
```

## Mortality

Import results data to list
```{r}
path = here("models_outputs/models_time/all", "mort", "table")
files = list.files(path)
 
all = lapply(files, function(x) get(assign(x, load(paste0(path, "/", x)))))
tudo <- bind_rows(all)
```

Remove intercept to include again as a separate column
```{r}
intercepts <- tudo %>% filter(term == "intercept")
names(intercepts)[8] = "intercept"

mortT <- tudo %>% filter(term != "intercept") %>%
  left_join(intercepts[,c(1:4,8)], by= c("sub","data", "fplot", "q_size")) %>%
  group_by(sub,data, fplot, q_size) %>% 
  mutate(VPC = variance/sum(variance))  %>% ungroup() %>%
   mutate(term = fct_relevel(term,"quadrat", "sp:time","time",  "quadrat:time",
                            "quadrat:sp", "sp", "Residual"),
         q_size = fct_recode(q_size, `5`= "quad_5", `10`= "quad_10",
                             `20`="quad_20",
                            `50`="quad_50",  `100`="quad_100")) %>%
  mutate(q_size = fct_relevel(q_size, "5","10", "20","50", "100")) %>%
  as.data.frame()

mortT$stdev <- mortT$Estimate
```

Number of mortality models per florest plot and quadrat scale.
```{r}
table(mortT$fplot, mortT$q_size)/7
```

## Recruitment

Import results data to list   
```{r}
path = here("models_outputs/models_time/all", "rec", "table")
files = list.files(path)
 
all = lapply(files, function(x) get(assign(x, load(paste0(path, "/", x)))))

tudo <- bind_rows(all)
```

Remove intercept to include again as a separate column
```{r}
intercepts <- tudo %>% filter(term == "intercept")
names(intercepts)[8] = "intercept"

recT <- tudo %>% filter(term != "intercept") %>%
  left_join(intercepts[,c(1:4,8)], by= c("sub","data","fplot", "q_size")) %>%
  group_by(sub,data, fplot, q_size) %>% 
  mutate(VPC = variance/sum(variance))  %>% ungroup() %>%
   mutate(term = fct_relevel(term,"quadrat", "sp:time", "time", "quadrat:time",
                            "quadrat:sp", "sp", "Residual"),
         q_size = fct_recode(q_size, `5`= "quad_5", `10`= "quad_10",
                             `20`="quad_20", `50` = "quad_50",
                             `100`="quad_100")) %>%
  mutate(q_size = fct_relevel(q_size, "5","10", "20","50", "100")) %>%
  as.data.frame()

recT$stdev <- recT$Estimate
```

Number of recruitment models per florest plot and quadrat scale.
```{r}
table(recT$fplot, recT$q_size)/7
```

## combining rates

All vital rates

```{r vitalT r2}
vitalT <- bind_rows(list(grow = growT, mort=mortT, rec=recT),.id="vital") 
```

To get ONE value per forest-vital, first I calculate the mean of the 10 medians of the posterior variances. Then I use this value as the Variance and take the VPC and R2.

```{r mvitalT mr2}
mvitalT <- vitalT %>% group_by(vital,data,fplot,q_size,term) %>%
  summarise(variance = mean(variance),
            stdev = mean(stdev)) %>%
   mutate(VPC = variance/sum(variance)) %>%
  mutate(term = fct_relevel(term, "quadrat", "sp:time","time", "quadrat:time", 
                            "quadrat:sp", "sp","Residual"))

mr2 <- mvitalT %>% filter(term != "Residual") %>%
  group_by(vital, data,fplot,q_size) %>% 
  summarise(r2 = sum(VPC))
```

Saving objects for models results
```{r}
save(mvitalT, vitalT, file= here("models_outputs", "all_results_time.Rdata"))
```


# All the 5ha subsampled plots

## SDs

```{r, fig.height=6, fig.width=10}
fsd <- vitalT %>% filter(q_size == "5") %>%
  ggplot(aes(x=fplot, y=stdev, col=term)) + 
  geom_quasirandom(alpha=0.5, groupOnX = TRUE) +
  facet_grid(vital~term) +
  stat_summary(fun=median, size=0.3, fatten=1,col="black",
                         geom="crossbar", width=0.7)+
  scale_color_manual(values = cores) +
  theme(legend.position = "none",
        panel.background = element_rect(colour="lightgray"),
        axis.text.x = element_text(angle=45,hjust=1)) +
  labs(title= "SD per subsample of 5ha - 5x5 quadrat scale")
fsd
```

```{r, fig.height=8, fig.width=10}
vitalT %>% filter(vital=="grow") %>%
  ggplot(aes(x=q_size, y=stdev, col=term)) + 
  geom_quasirandom(alpha=0.5, groupOnX = TRUE) +
  facet_grid(fplot~term, labeller=labplot) +
  
  stat_summary(fun=median, size=0.3, fatten=1,col="black",
                         geom="crossbar", width=0.7)+
  scale_color_manual(values = cores) +
    xlab("Spatial scale") +
  theme(legend.position = "none",
        panel.background = element_rect(colour="lightgray"),
        axis.text.x = element_text(angle=45,hjust=1)) +
  labs(title= "SD per subsample of 5ha - GROWTH")
```

```{r, fig.height=8, fig.width=10}
vitalT %>% filter(vital=="mort") %>%
  ggplot(aes(x=q_size, y=stdev, col=term)) + 
  geom_quasirandom(alpha=0.5, groupOnX = TRUE) +
  facet_grid(fplot~term, labeller=labplot) +
  
  stat_summary(fun=median, size=0.3, fatten=1,col="black",
                         geom="crossbar", width=0.7)+
  scale_color_manual(values = cores) +
    xlab("Spatial scale") +
  theme(legend.position = "none",
        panel.background = element_rect(colour="lightgray"),
        axis.text.x = element_text(angle=45,hjust=1)) +
  labs(title= "SD per subsample of 5ha - MORTALITY")
```

```{r, fig.height=8, fig.width=10}
vitalT %>% filter(vital=="rec") %>%
  ggplot(aes(x=q_size, y=stdev, col=term)) + 
  geom_quasirandom(alpha=0.5, groupOnX = TRUE) +
  facet_grid(fplot~term, labeller=labplot) +
  
  stat_summary(fun=median, size=0.3, fatten=1,col="black",
                         geom="crossbar", width=0.7)+
  scale_color_manual(values = cores) +
    xlab("Spatial scale") +
  theme(legend.position = "none",
        panel.background = element_rect(colour="lightgray"),
        axis.text.x = element_text(angle=45,hjust=1)) +
  labs(title= "SD per subsample of 5ha - RECRUITMENT")
```

## VPCs

```{r, fig.height=6, fig.width=10}
fvpc <- vitalT %>% filter(q_size =="5") %>%
  ggplot(aes(x=fplot, y=VPC, col=term)) + 
  geom_quasirandom(alpha=0.5) +
  facet_grid(vital~term, labeller= grpvitT ) +
  stat_summary(fun=median, size=0.3, fatten=1,col="black",
                         geom="crossbar", width=0.7)+
  scale_color_manual(values = cores) +
  theme(legend.position = "none",
                panel.background = element_rect(colour="lightgray"),
        axis.text.x = element_text(angle=45,hjust=1)) +
  labs(title= "VPC per subsample of 5ha - 5x5")
fvpc
```

```{r, fig.height=8, fig.width=10}
fig <- vitalT %>% filter(vital=="grow") %>%
  ggplot(aes(x=q_size, y=VPC, col=term)) + 
  geom_quasirandom(alpha=0.5, groupOnX = TRUE) +
  facet_grid(fplot~term, labeller=labplot) +
  
  stat_summary(fun=median, size=0.3, fatten=1,col="black",
                         geom="crossbar", width=0.7)+
  scale_color_manual(values = cores) +
    xlab("Spatial scale") +
  theme(legend.position = "none",
        panel.background = element_rect(colour="lightgray"),
        axis.text.x = element_text(angle=45,hjust=1)) +
  labs(title= "VPC per subsample of 5ha - GROWTH")
fig
```

```{r, fig.height=8, fig.width=10}
fig <- vitalT %>% filter(vital=="mort") %>%
  ggplot(aes(x=q_size, y=VPC, col=term)) + 
  geom_quasirandom(alpha=0.5, groupOnX = TRUE) +
  facet_grid(fplot~term, labeller=labplot) +
  
  stat_summary(fun=median, size=0.3, fatten=1,col="black",
                         geom="crossbar", width=0.7)+
  scale_color_manual(values = cores) +
    xlab("Spatial scale") +
  theme(legend.position = "none",
        panel.background = element_rect(colour="lightgray"),
        axis.text.x = element_text(angle=45,hjust=1)) +
  labs(title= "VPC per subsample of 5ha - MORTALITY")
fig
```

```{r, fig.height=8, fig.width=10}
fig <- vitalT %>% filter(vital=="rec") %>%
  ggplot(aes(x=q_size, y=VPC, col=term)) + 
  geom_quasirandom(alpha=0.5, groupOnX = TRUE) +
  facet_grid(fplot~term, labeller=labplot) +
  
  stat_summary(fun=median, size=0.3, fatten=1,col="black",
                         geom="crossbar", width=0.7)+
  scale_color_manual(values = cores) +
    xlab("Spatial scale") +
  theme(legend.position = "none",
        panel.background = element_rect(colour="lightgray"),
        axis.text.x = element_text(angle=45,hjust=1)) +
  labs(title= "VPC per subsample of 5ha - RECRUITMENT")
fig
```


# Average per forest

## SD stacked per forest

5x5 m

```{r}
mvitalT %>% filter(q_size == "5") %>%
ggplot(aes(x=fplot, y=stdev, fill=term, col=term))+
  geom_col(alpha=0.7)+
  scale_color_manual(values = cores) +
  scale_fill_manual(values = cores)+
  facet_grid(~vital)
```

## VPC per forest

5x5 m

```{r}
ah <- mvitalT %>% filter(q_size == "5") %>% group_by(vital,term) %>%
  summarise(VPC_mean = as.character(round(mean(VPC), 2))) %>%
  mutate(x_val = as.numeric(VPC_mean), 
         y_val = 0.6:6.8)
ah[c(11,13),3] <-  c("0.10", "0.30")
```

```{r}
pm<- mvitalT %>% filter(q_size == "5") %>%
  ggplot(aes(y=rev(term), x=VPC)) +
  facet_grid(~vital, labeller = grpvitT, scales="free_x") +
  geom_jitter(size=5, alpha=0.5, aes(shape=fplot, fill=term), width = 0,
              height = 0.15, col="black") +
  stat_summary(fun=mean, fun.min=mean, fun.max=mean,
               geom="crossbar", width=0.6, fatten=3) +
  scale_fill_manual(name= "Forest", values=cores, guide ="none")+
  scale_shape_manual(name="Forest",values=c(22,21,24,23,25), #c(15,16,17,18,8)
                     labels= c("Barro Colorado", "Fushan", "Lambir","Luquillo",
                               "Pasoh", "Wytham Woods")) +
  geom_vline(xintercept=c(0,0.2,0.4,0.6), col='gray', linetype="dashed",
             size=0.2) +
  scale_y_discrete(labels=c("Residual", "Species", "Species x Space", 
                                "Space x Time", "Time", "Species x Time",
                                "Space"))+
  theme(legend.position= "bottom",
        plot.margin =  margin(0.1, 0.1, 0.1, 0, "cm"),
        legend.background = element_rect(fill="gray95"),
        panel.spacing.y = unit(0.8, "cm"),
        panel.background = element_rect(colour="lightgray"),
        text = element_text(size=16),
        axis.text = element_text(size=14)) +
  ylab("") + xlab("VPC") +
  geom_text(data=ah, aes(x=x_val, y=rev(y_val), label=VPC_mean), hjust=0,
            size=4)
pm
```

```{r}
mvitalT.sum <-  mvitalT %>% filter(q_size=="5") %>% ungroup()%>%
  group_by(vital,term) %>%
  summarise(meanVPC = mean(VPC),
            sdVPC = sd(VPC)) %>%
  pivot_wider(names_from = vital, values_from = c(meanVPC, sdVPC)) %>%
  mutate_if(is.numeric, round, digits=2) %>%
  select(term, meanVPC_grow, sdVPC_grow, meanVPC_mort, sdVPC_mort,
         meanVPC_rec, sdVPC_rec)


mvitalT.sum %>%
  htmlTable(header=c("term" ,rep(c("Mean", "SD"), 3)),
            cgroup = c("", "Growth", "Mortality", "Recruitment"),
            n.cgroup =c(1,2,2,2))

```

# TOTAL VPC per vital rate 5x5

To make the bar plots, first I sum all the variances per term and then calculate VPC from this sum

```{r}
tot.vpc.fig <- mvitalT %>% filter(q_size=="5") %>% ungroup()%>%
  group_by(vital,term) %>%
  summarise(sumVar =sum(variance)) %>%
  mutate(VPC = sumVar/sum(sumVar)) %>%
ggplot(aes(x=vital, fill=term, col=term, y=VPC)) + 
  geom_col(alpha=0.7, width = 0.6) +
  facet_grid(~vital, scales="free_x", labeller = labvit) +
   xlab("") + ylab("Mean VPC") +
    scale_fill_manual(values=cores, labels=grpTlab) +
  scale_color_manual(values=cores, labels=grpTlab) +
  theme(legend.position = "none",
         panel.spacing.x = unit(0.01, "cm"),
        axis.text.x = element_blank(),
        axis.ticks = element_blank(),
            text = element_text(size=16),
        axis.text = element_text(size=14))
tot.vpc.fig
```

# FIGURE 2

```{r fig2}
pm +  labs(tag="(a)") +
  theme( plot.margin =  margin(0.1, 0.1, 0.1, 0, "cm"),
         plot.tag.position = c(0,0.93),
         plot.tag = element_text(size=20, vjust = -0.6, hjust = -1)) +
tot.vpc.fig + labs(tag="(b)") +
    theme(plot.tag.position = c(0,0.93),
         plot.tag = element_text(size=20, vjust = -0.6, hjust = 0.1)) +
    plot_layout(widths = c(1,0.4))
ggsave(here("figures","FIG_2.jpeg"), width=14, height=6)
```

# FIGURE 3 VCP bar all scales

```{r}
mvitalT %>% group_by(vital,q_size,term) %>%
  summarise(sumVar =sum(variance)) %>%
  group_by(vital,q_size)%>%
  mutate(VPC = sumVar/sum(sumVar)) %>% select(-sumVar) %>%
  pivot_wider(names_from =q_size, values_from = VPC) %>%
  mutate_if(is.numeric, round, digits=2) %>%
  htmlTable()
```

```{r}
barvpc <- mvitalT %>% group_by(vital,q_size,term) %>%
  summarise(sumVar =sum(variance)) %>%
  group_by(vital,q_size)%>%
  mutate(VPC = sumVar/sum(sumVar)) %>%
  ggplot(aes(x=q_size, y=VPC, fill=term, col=term))+
  geom_col(alpha=0.7)+
  scale_color_manual(values = cores, labels=grpTlab) +
  scale_fill_manual(values = cores, labels=grpTlab )+
  facet_grid(~vital,labeller = grpvitT) +
  xlab("Spatial grain (quadrat size)") +
  ylab("VPC") + labs(tag="(b)")
barvpc
png(here("figures", "FIG_3_VPC_time_scales.png"), height = 550, width=1300, res=130)
barvpc
dev.off() 
```
