---
title: "Rare species patterns and results excluding rare species for no-time models"
author: "Melina Leite"
date: "`r  Sys.Date()`"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "..") })
output:
  rmdformats::readthedown:
  highlight: kate
self_contained: true
thumbnails: false
lightbox: true
gallery: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = T, warning = F, fig.align = "center", message = F,
  error = F, cache = F, cache.lazy = FALSE)

library(knitr)
library(tidyverse); library(cowplot); library(patchwork); library(ggrepel)
theme_set(theme_cowplot())
library(ggbeeswarm)
library(DirichletReg)
library(here)
library(htmlTable)
```


# Classifying rare species

Species classification was done for each forest plot and time interval separately. We used the living and dead individuals of the final census from the interval. We used the FuzzyQ clustering algorithm in the ‘FuzzyQ’ R package (Balbuena et al., 2021) to estimate the probability of each species to be common or rare based on species abundance and occupancy in 50x50 m quadrats.

To describe and compare rarity patterns in all forest, we estimated the number of species, number of individuals, and density of rare and common species. For the forests plots with more than 1 census interval, we averaged the estimates across census intervals. 

Here, we provide the list of species code and the classification in a Rdata file.

```{r}
load(here("data/rare_species_classification.Rdata"))

# plot variables including
plots_info <- read.table(here("data/plot_sites_information.csv"), sep=",", header=T)

classi <-classi %>% left_join(plots_info[,c(1,6,9,12,17:23,28)], by=c("fplot" = "ID"))
classi$fplot.time = paste0(classi$fplot,"_", classi$time)
```



```{r}
rich = classi %>%  count(fplot.time,rarity) %>% rename(rich=n)
ntree = classi %>% group_by(fplot.time,rarity) %>% 
  summarise(ntrees = sum(n))

dat <- rich %>% left_join(ntree, by=c("fplot.time", "rarity")) %>%
  pivot_wider(names_from = rarity, values_from =c(rich, ntrees)) %>%
  mutate(p.rich_common = rich_common*100/(rich_common+rich_rare),
         p.rich_rare = rich_rare*100/(rich_common+rich_rare),
         p.ntrees_common = ntrees_common*100/(ntrees_common+ntrees_rare),
         p.ntrees_rare = ntrees_rare*100/(ntrees_common+ntrees_rare))

#Mean per forest plots with more than 1 census interval
dat.m <- dat %>% separate(fplot.time, c("fplot", "time")) %>%
  group_by(fplot) %>% 
  summarise_at(vars(rich_common:p.ntrees_rare),mean) %>%
  group_by(fplot) %>% 
  summarise_at(vars(rich_common:p.ntrees_rare),round,digits=0) %>%
  select(fplot,rich_common, p.rich_common, rich_rare, p.rich_rare,
         ntrees_common, p.ntrees_common, ntrees_rare, p.ntrees_rare)
```

```{r}
dat.m %>%
  htmlTable(header=c("Forest" ,rep(c("N", "%"), 4)),
            cgroup = list(c("", "Species richness", "Number of trees"),
                           c("", "Common", "Rare", "Common", "rare")),
            n.cgroup = list(c(1,4,4),
                             c(1,2,2,2,2)),caption="Table S4.1. Number and percentages of species and trees classified as common or rare per forest plot. See Table S1.1 and S1.2 for forest plot names and information. For forest plots with more than 1 census interval, we average the values across intervals."
            )
```

```{r tableS4.2}
classi %>% group_by(rarity) %>%
  mutate(dens.ha = n/size_data) %>%
  summarise(minD = min(dens.ha),
            maxD = max(dens.ha),
            meanD = mean(dens.ha),
            sdD = sd(dens.ha), 
            medianD = median(dens.ha),
            q90D = quantile(dens.ha, 0.9),
            q95D = quantile(dens.ha, 0.95)) %>%
  kable(digits=2, caption = "Table S4.2. Summary values for the density of trees (N/ha) classified as common or rare for the 21 forest plots. SD - standard deviation, Quant - quantiles.")
```



```{r}
dat.m2 <- dat.m %>% 
  left_join(plots_info[,c(1,6,9,12,17:23,28)], by=c("fplot" = "ID")) %>%
  mutate(tot.rich = rich_common + rich_rare,
         tot.n = ntrees_common+ntrees_rare)
```

```{r, fig.height=10, fig.cap="Figure S4.1. Percentages of rare and common species and number of trees per forest plot. Forest plots are arranged by absolute latitude (from the largest to the smallest). See Table S1.2 for plots names."}
prich <- dat.m2 %>%
  pivot_longer(c(p.rich_common, p.rich_rare),names_to = "rarity",
               values_to = "richness") %>%
  ggplot(aes(x=fct_reorder(fplot, tot.rich, max),
             y=richness, fill=rarity)) +
    geom_col() +
  xlab("") + ylab("Species richness (%)") +
  labs(tag="A")+
  scale_fill_discrete("Species",
                      labels=c("common", "rare"))+
  theme(axis.text.x = element_text(angle = 45, hjust=1))

p.ntree <- dat.m2 %>%
  pivot_longer(c(p.ntrees_common, p.ntrees_rare),names_to = "rarity",
               values_to = "ntrees") %>%
  ggplot(aes(x=fct_reorder(fplot, Nsp_literature, max),
             y=ntrees, fill=rarity)) +
    geom_col()+
  labs(tag="B")+
    xlab("") + ylab("Number of trees (%)") +

  scale_fill_discrete("Species",
                      labels=c("common", "rare"))+
  theme(axis.text.x = element_text(angle = 45, hjust=1),
        legend.position = "none")

prich/p.ntree + plot_layout(guides = "collect")
ggsave(here("figures/FIG_S4.1_class_species.jpeg"), height = 7, width = 6, unit="in")
```






```{r}
library(wesanderson) #better colour palette
# Gradient color for latitud
pal <- wes_palette("Zissou1",21, type = "continuous")[21:1] # azul frio-temperado
```

```{r, fig.cap="Figure S4.2. Percentage of rare species against the number of species (log10 scale) for 21 worldwide distributed forest plots. Blue line and grey area are the fitted results and confidence intervals for a generalised additive model showing a decrease in the percentage of rare species with the number of species but only for forests with less than 100 species. Model's adjusted R2 = 0.30. Each forest plot is coloured by the latitude in absolute values."}
dat.m2 %>% 
  ggplot(aes(x=tot.rich, y=p.rich_rare))+ 
  scale_x_log10()+
  geom_point(aes( col=abs(lat)), size=3) +
  scale_color_gradientn(name="Latitude \n (abs)",colors=pal) +
  geom_smooth(method=mgcv::gam,se=T, formula = y ~ s(x, bs = "cs")) +
  xlab("Number of species") +
  ylab("Percentage of rare species") 
ggsave(here("figures/FIG_S4.2_propRare_nsp.jpg"), width=8, height = 5)
```

```{r}
mod <- mgcv::gam(p.rich_rare ~ s(log(tot.rich), bs="cs"), data=dat.m2, method = "REML")
summary(mod)
```



# No time models

Running models with different datasets:

-   **all**: original data, all species all individuals

-   **exclude**: exclude rare species

-   **regroup**: regrouping observations of rare species in ONE species group

Using only the **5x5 m quadrat scale**

```{r run}
#Figures colors/labels
cores <- c("#c00000", "#b666d2", "#70ad47", "#ffc000")

labvit <- as_labeller(c(grow = "Growth", mort="Mortality", 
                        rec="Recruitment"))
grplab = as_labeller(c(quadrat = "Space", `quadrat:sp`= "Space X Species",
                       sp="Species", Residual = "residual"))
grpvit <- as_labeller(c(grow = "Growth", mort="Mortality", 
                        rec="Recruitment",
                        quadrat = "Space", `quadrat:sp`= "Species X Space",
                        sp="Species", Residual = "Residual"))
```

## Data

Growth

```{r}
grp <- c("all","exclude", "regroup")
tudo <- list()
path1 = here("models_outputs/models_notime/")
for (i in 1:length(grp)){
  path <- paste0(path1, grp[i], "/grow/table")
  files = list.files(path)
  all <-  lapply(files, function(x) get(assign(x, load(paste0(path, "/", x)))))
  names(all) <- substr(files,1,nchar(files)-6)
  tudo[[i]] <- all
}
names(tudo) <- grp
tes <- map(tudo,bind_rows, .id="id") %>% bind_rows() %>% filter(q_size=="quad_5")

#tes %>% count(data,fplot) %>% mutate(n=n/8)
table(tes$fplot,tes$data)/5

#remove intercept to include again as a separate column
intercepts <- tes %>% filter(term == "intercept")
names(intercepts)[9] = "intercept"

grow <- tes %>% filter(term != "intercept") %>%
  left_join(intercepts[,c(1:4,9)], by= c("id", "data", "fplot", "time")) %>%
  group_by(id, data, fplot, time) %>% 
  mutate(VPC = variance/sum(variance))  %>% ungroup() %>%
  mutate(term = fct_relevel(term, "quadrat", "quadrat:sp", "sp", "Residual")) %>%
  as.data.frame()

# divide the sd of the terms by the mean growth (intercept) for each dataset.
grow$stdev = grow$Estimate/grow$intercept
```

Mort

```{r}
grp <- c("all","exclude", "regroup")
tudo <- list()
path1 = here("models_outputs/models_notime/")
for (i in 1:length(grp)){
  path <- paste0(path1,grp[i], "/mort/table")
  files = list.files(path)
  all <-  lapply(files, function(x) get(assign(x, load(paste0(path, "/", x)))))
  names(all) <- substr(files,1,nchar(files)-6)
  tudo[[i]] <- all
}
names(tudo) <- grp
tes <- map(tudo,bind_rows, .id="id") %>% bind_rows() %>% filter(q_size=="quad_5")

#tes %>% count(data,fplot) %>% mutate(n=n/8)
table(tes$fplot,tes$data)/5

#remove intercept to include again as a separate column
intercepts <- tes %>% filter(term == "intercept")
names(intercepts)[9] = "intercept"

mort <- tes %>% filter(term != "intercept") %>%
  left_join(intercepts[,c(1:4,9)], by= c("id", "data", "fplot", "time")) %>%
  group_by(id, data, fplot, time) %>% 
  mutate(VPC = variance/sum(variance))  %>% ungroup() %>%
  mutate(term = fct_relevel(term, "quadrat", "quadrat:sp", "sp", "Residual")) %>%
  as.data.frame()
mort$stdev = mort$Estimate
```

Rec

```{r}
grp <- c("all","exclude", "regroup")
tudo <- list()
path1 = here("models_outputs/models_notime/")
for (i in 1:length(grp)){
  path <- paste0(path1,grp[i], "/rec/table")
  files = list.files(path)
  all <-  lapply(files, function(x) get(assign(x, load(paste0(path, "/", x)))))
  names(all) <- substr(files,1,nchar(files)-6)
  tudo[[i]] <- all
}
names(tudo) <- grp
tes <- map(tudo,bind_rows, .id="id") %>% bind_rows() %>% filter(q_size=="quad_5")

#tes %>% count(data,fplot) %>% mutate(n=n/8)
table(tes$fplot,tes$data)/5

#remove intercept to include again as a separate column
intercepts <- tes %>% filter(term == "intercept")
names(intercepts)[9] = "intercept"

rec <- tes %>% filter(term != "intercept") %>%
  left_join(intercepts[,c(1:4,9)], by= c("id", "data", "fplot", "time")) %>%
  group_by(id, data, fplot, time) %>% 
  mutate(VPC = variance/sum(variance))  %>% ungroup() %>%
  mutate(term = fct_relevel(term, "quadrat", "quadrat:sp", "sp", "Residual")) %>%
  as.data.frame()
rec$stdev = rec$Estimate
```

Excluding recruitment for Wytham Woods, as in "all data".
```{r}
rec <- rec %>% filter(fplot!="wyw")
```


Combining results.
```{r}
comb <- bind_rows(list(grow=grow, mort=mort, rec=rec), .id="vital") %>%
  mutate(term = fct_relevel(term, "quadrat", "quadrat:sp", "sp","Residual")) %>%
  ungroup() 
```




## Comparing groups

```{r, fig.height=10, fig.width=20}
comb %>% 
  ggplot(aes(x=data, y=VPC)) +
  geom_col(aes(fill=term, color=term), alpha=0.7) +
  scale_fill_manual(values=cores) +
  scale_color_manual(values=cores) +
  facet_grid(vital~fplot+time) +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

mean over fplots

```{r}
comb %>% group_by(vital,data,term) %>% summarise(VPC=mean(VPC)) %>%
  ggplot(aes(x=data, y=VPC)) +geom_col(aes(fill=term, color=term), alpha=0.7) +
  scale_fill_manual(values=cores) + 
  scale_color_manual(values=cores) +
  facet_grid(~vital)+
  theme(axis.text.x = element_text(angle=45, hjust=1))+
  labs(tag="a)")
#ggsave("figs/FIG_S4.4a_groups_mean.jpg", width=20, height = 10, units = "cm")
```

```{r, fig.height=8}
comb %>%
  ggplot(aes(x=data,y=VPC, col=data)) + geom_boxplot()+
  facet_grid(vital~term) +
  theme(axis.text.x = element_text(angle=45,hjust=1),
        legend.position = "none")
```

```{r}
comb %>% mutate(xis = as.numeric(as.factor(data))) %>%
  ggplot(aes(x=xis,y=stdev, col=paste(fplot,time))) + geom_line() +
  facet_grid(vital~term) + 
  scale_x_continuous(breaks = 1:3, labels=levels(comb$data),
                     name="") +
  theme(axis.text.x = element_text(angle=45, hjust=1),
        legend.position = "none")
```

```{r}
comb %>% mutate(xis = as.numeric(as.factor(data))) %>%
  ggplot(aes(x=xis,y=VPC, col=paste(fplot,time))) + geom_line() +
  facet_grid(vital~term) + 
  scale_x_continuous(breaks = 1:3, labels=levels(comb$data),
                     name="") +
  theme(axis.text.x = element_text(angle=45, hjust=1),
        legend.position = "none")
```



## Difference to all data

### Absolute difference

```{r}
dif <- comb %>% 
  pivot_wider(id_cols = c(vital,fplot,time,term),
              names_from = data,
              values_from = c(stdev,VPC, richness)) %>%
  mutate(stdev.dif_exclude = stdev_exclude - stdev_all,
         VPC.dif_exclude =  VPC_exclude - VPC_all,
         stdev.dif_regroup = stdev_regroup - stdev_all,
         VPC.dif_regroup =  VPC_regroup - VPC_all) %>%
  select(vital,fplot,time,term,richness_exclude,
         stdev.dif_exclude,
         stdev.dif_regroup, VPC.dif_exclude, VPC.dif_regroup) %>%
  pivot_longer(6:9, names_to = c("index", "data"), names_sep = "_",
                #names_pattern = "(.*_).(.*)",
                values_to =  "value") %>%
  pivot_wider(names_from = index, values_from = value)
```

```{r, fig.height=6, fig.width=8}
dif %>%
  ggplot(aes(x=data, y=stdev.dif)) + 
  facet_grid(vital~term, scales= "free_x") +
  geom_boxplot() +
  geom_quasirandom(aes(col=fplot)) +
  ylab("SD no.rare - SD all") + xlab("") +
  geom_hline(yintercept = 0, linetype= "dotted")  +
  ggtitle("Difference in SD") +
  theme(legend.position = "right",
        panel.background = element_rect(colour="lightgray"),
        axis.text.x = element_text(angle=45, hjust=1))
```

```{r, fig.height=6, fig.width=8}
dif %>%
  ggplot(aes(x=data, y=VPC.dif)) + 
  facet_grid(vital~term, scales= "free_x") +
  geom_boxplot() +
  geom_quasirandom(aes(col=fplot)) +
  ylab("VPC no.rare - VPC all") + xlab("") +
  geom_hline(yintercept = 0, linetype= "dotted")  +
  ggtitle("Difference in VPC") +
  theme(legend.position = "right",
        panel.background = element_rect(colour="lightgray"),
        axis.text.x = element_text(angle=45, hjust=1))
```

### Relative difference SD

```{r}
difr <- comb %>% 
  pivot_wider(id_cols = c(vital,fplot,time,term),
              names_from = data,
              values_from = c(stdev,VPC, richness)) %>%
  mutate(stdev.dif_exclude = (stdev_exclude - stdev_all)*100/stdev_all,
         stdev.dif_regroup = (stdev_regroup - stdev_all)*100/stdev_all) %>%
  select(vital,fplot,time,term,richness_exclude,
         stdev.dif_exclude,stdev.dif_regroup) %>%
  pivot_longer(6:7, names_to = "data",
                values_to =  "stdev.dif") %>%
  mutate(data = substr(data,11, nchar(data)))
```

```{r, fig.height=6, fig.width=8}
difr %>%
  ggplot(aes(x=data, y=stdev.dif)) + 
  facet_grid(vital~term, scales= "free_x") +
  geom_boxplot() +
  geom_quasirandom(aes(col=fplot)) +
  ylab("Proportional difference (%)") + xlab("") +
  geom_hline(yintercept = 0, linetype= "dotted")  +
  ggtitle("Relative difference in SD") +
  theme(legend.position = "right",
        panel.background = element_rect(colour="lightgray"),
        axis.text.x = element_text(angle=45, hjust=1))
```

```{r}
difr %>% group_by(vital, term, data) %>% summarise(mean=mean(stdev.dif, na.rm=T)) %>%
  pivot_wider(names_from = term, values_from = mean)
```

#### table

Relative differences in SD

```{r}
difr %>% group_by(vital, term, data) %>% 
  summarise(mean=mean(stdev.dif, na.rm=T)) %>%
  pivot_wider(names_from = c(vital,data), values_from = mean) %>%
  kable(digits=1)
```

Mean absolute differences in VPC

```{r}
dif %>% group_by(vital, term, data) %>% 
  summarise(mean=mean(VPC.dif, na.rm=T)) %>%
  pivot_wider(names_from = c(vital,data), values_from = mean) -> write
kable(write)
```

## Comparing with NUMBER of rare species

```{r, fig.height=6, fig.width=12}
dif %>%
  ggplot(aes(x=richness_exclude, y=stdev.dif, shape=data)) + 
  geom_point(aes(col=fplot)) + 
  geom_smooth(method="lm", se=F, aes(linetype=data)) +
  facet_grid(vital~term) +
  ylab("Difference in SD to original data") +
  xlab("log N of rare species") +
  geom_hline(yintercept = 0, linetype= "dotted")  +
  ggtitle("Difference in SD") +
  scale_x_log10()+
  theme(axis.text.x = element_text(angle=45, hjust=1),
        )
```

```{r, fig.height=6, fig.width=12}
difr %>%
  ggplot(aes(x=richness_exclude, y=stdev.dif, shape=data)) + 
  geom_point(aes(col=fplot)) + 
  geom_smooth(method="lm", se=F, aes(linetype=data)) +
  facet_grid(vital~term) +
  ylab("Relative difference in SD to original data") +
  xlab("log N of rare species") +
  geom_hline(yintercept = 0, linetype= "dotted")  +
  ggtitle("Proportional difference in SD to original data") +
  theme(axis.text.x = element_text(angle=45, hjust=1),
        panel.background = element_rect(fill="white", color="black"))
```

```{r, fig.height=6, fig.width=12}
pm <- dif %>%
  ggplot(aes(x=richness_exclude, y=VPC.dif, shape=data)) + 
  geom_point(aes(col=fplot)) + 
  facet_grid(vital~term) +
   geom_smooth(method="lm", se=F, aes(linetype=data)) +
  ylab(" Absolute difference in VPC to original data") +
   xlab("log N of rare species") +
  geom_hline(yintercept = 0, linetype= "dotted")  +
  theme(axis.text.x = element_text(angle=45, hjust=1),
        panel.background = element_rect(fill="white", color="black"))
pm
```


## Dirichlet regression for exclude results with species richness

Richness by rarefaction at the 6ha plot size (sampling unit 20x20m)

```{r}
load(here("data", "rarefaction.curves.Rdata"))
rich.rare <- rare[rare$sites ==150,2:3]
colnames(rich.rare)[1] <- "richness.rarefaction"
comb <- comb %>% left_join(rich.rare, by="fplot")
```

Latitude

```{r}
load(here("data", "plots_structure.Rdata"))
comb <- comb %>% left_join(plots.structure[,1:2], "fplot")
```

```{r}
#mean for each forest
mcomb <- comb %>% group_by(vital,data, fplot,term, lat) %>%
  summarise(VPC = mean(VPC),
            stdev = mean(stdev),
            richness.rarefaction = mean(richness.rarefaction))
```

### GROWTH

```{r}
predirig <-  mcomb %>% 
  filter(vital == "grow", data == "exclude") %>%
  select(fplot, term, VPC, richness.rarefaction,lat) 
diridatag <- predirig %>%
  pivot_wider(names_from = term, values_from = VPC) %>% ungroup() %>%
  mutate(log.rich = log(richness.rarefaction),
         log.rich.o=log.rich) %>%
  mutate_at(vars( log.rich), scale)

vpcg <- DR_data(diridatag[,c("quadrat", "sp", "quadrat:sp",  "Residual")])
#plot(vpc)
```

```{r, echo=T}
grow2 <- DirichReg(vpcg~log.rich, data=diridatag, model="alternative", base=4)
summary(grow2)
```

```{r}
newdata <- expand.grid(log.rich = seq(min(diridatag$log.rich),
                                      max(diridatag$log.rich), length.out=10))
pred <- predict(grow2, newdata = newdata, se=T)
confint(grow2)
colnames(pred) <- colnames(vpcg)
newdata$log.rich.o <- newdata$log.rich*sd(diridatag$log.rich.o) +
  mean(diridatag$log.rich.o)
newdata$rich.o <- exp(newdata$log.rich.o)
newpredg <- cbind(newdata,pred) %>% pivot_longer(`quadrat`:`Residual`, names_to="term", values_to="pred")

newpredg %>% mutate(term = fct_relevel(term, "quadrat", "quadrat:sp", "sp",
                                       "Residual")) %>%
  ggplot(aes(x=rich.o, y=pred, col=term)) +
  geom_line() + facet_grid(~term) +
  geom_point(data=predirig, aes(x=richness.rarefaction, y=VPC, col=term)) +
  scale_x_log10() +
  ggtitle("Mod log.rich")
```

Residuals

```{r}
resid <- residuals(grow2, type = "standardized")
resid <-  data.frame(resid[,1:4])
resid$log.rich <- diridatag$log.rich
resid <- resid %>% pivot_longer(1:4, names_to="term", values_to="resid")

ggplot(resid, aes(x=log.rich, y=resid)) + geom_point() +
  facet_grid(~term)
```

### MORTALITY

```{r}
predirim <-  mcomb %>% 
  filter(vital == "mort", data == "exclude") %>%
  select(fplot, term, VPC, richness.rarefaction,lat) 
diridatam <- predirim %>%
  pivot_wider(names_from = term, values_from = VPC) %>% ungroup() %>%
  mutate(log.rich = log(richness.rarefaction),
         log.rich.o=log.rich) %>%
  mutate_at(vars( log.rich), scale)

vpcm <- DR_data(diridatam[,c("quadrat", "sp", "quadrat:sp",  "Residual")])
#plot(vpc)
```

```{r, echo=T}
mort2 <- DirichReg(vpcm~log.rich,data=diridatam, model="alternative", base=4)
summary(mort2)
```

```{r}
newdata <- expand.grid(log.rich = seq(min(diridatam$log.rich),
                                      max(diridatam$log.rich), length.out=10))
pred <- predict(mort2, newdata = newdata, se=T)
confint(mort2)
colnames(pred) <- colnames(vpcm)
newdata$log.rich.o <- newdata$log.rich*sd(diridatam$log.rich.o) +
  mean(diridatam$log.rich.o)
newdata$rich.o <- exp(newdata$log.rich.o)
newpredm <- cbind(newdata,pred) %>% pivot_longer(`quadrat`:`Residual`, names_to="term", values_to="pred")

newpredm %>% mutate(term = fct_relevel(term, "quadrat", "quadrat:sp", "sp",
                                       "Residual")) %>%
  ggplot(aes(x=rich.o, y=pred, col=term)) +
  geom_line() + facet_grid(~term) +
  geom_point(data=predirim, aes(x=richness.rarefaction, y=VPC, col=term)) +
  scale_x_log10() +
  ggtitle("Mod log.rich")
```

Residuals
```{r}
resid <- residuals(mort2, type = "standardized")
resid <-  data.frame(resid[,1:4])
resid$log.rich <- diridatam$log.rich
resid <- resid %>% pivot_longer(1:4, names_to="term", values_to="resid")

ggplot(resid, aes(x=log.rich, y=resid)) + geom_point() +
  facet_grid(~term)
```

### RECRUITMENT

```{r}
predirir <-  mcomb %>% 
  filter(vital == "rec", data == "exclude") %>%
  select(fplot, term, VPC, richness.rarefaction,lat) 
diridatar <- predirir %>%
  pivot_wider(names_from = term, values_from = VPC) %>% ungroup() %>%
  mutate(log.rich = log(richness.rarefaction),
         log.rich.o=log.rich) %>%
  mutate_at(vars( log.rich), scale)

vpcr <- DR_data(diridatar[,c("quadrat", "sp", "quadrat:sp",  "Residual")])
#plot(vpc)
```

```{r}
rec2 <- DirichReg(vpcr~log.rich,data=diridatar, model="alternative", base=4)
summary(rec2)
```

```{r}
newdata <- expand.grid(log.rich = seq(min(diridatar$log.rich),
                                      max(diridatar$log.rich), length.out=10))
pred <- predict(rec2, newdata = newdata, se=T)
colnames(pred) <- colnames(vpcr)
newdata$log.rich.o <- newdata$log.rich*sd(diridatar$log.rich.o) +
  mean(diridatar$log.rich.o)
newdata$rich.o <- exp(newdata$log.rich.o)
newpredr <- cbind(newdata,pred) %>% pivot_longer(`quadrat`:`Residual`, names_to="term", values_to="pred")

newpredr %>% mutate(term = fct_relevel(term, "quadrat", "quadrat:sp", "sp",
                                       "Residual")) %>%
  ggplot(aes(x=rich.o, y=pred, col=term)) +
  geom_line() + facet_grid(~term) +
  geom_point(data=predirir, aes(x=richness.rarefaction, y=VPC, col=term)) +
  scale_x_log10() +
  ggtitle("Mod log.rich")
```

Residuals

```{r}
resid <- residuals(rec2, type = "standardized")
resid <-  data.frame(resid[,1:4])
resid$log.rich <- diridatar$log.rich
resid <- resid %>% pivot_longer(1:4, names_to="term", values_to="resid")

ggplot(resid, aes(x=log.rich, y=resid)) + geom_point() +
  facet_grid(~term)
```

### FIGURE

Calculated prediction interval

```{r}
source(here("scripts/prediction_intervals_dirichlet_exclude.R"), local=T)
#load("results/prediction_intervals_dirichlet_exclude.Rdata") #  quants
```

```{r}
predis <- bind_rows(list(grow=newpredg, mort=newpredm, rec=newpredr),
                    .id="vital") %>% 
  mutate(term = fct_relevel(term, "quadrat", "quadrat:sp", "sp","Residual"))

pontos <- bind_rows(list(grow=predirig, mort= predirim, rec=predirir), .id="vital")%>% 
  mutate(term = fct_relevel(term, "quadrat", "quadrat:sp", "sp", "Residual"))
quants <- bind_rows(list(grow=quantg,mort=quantm, rec=quantr), .id="vital")%>% 
  mutate(term = fct_relevel(term, "quadrat", "quadrat:sp", "sp", "Residual"))
```

```{r}
# pvalues
res <- data.frame(vital = c("grow", "mort", "rec"),
                  P = NA,
                  term = "Residual")

test <- summary(grow2)
vals <- as.data.frame(test$coef.mat)[1:6,]
vals$coef <- names(test$coefficients)[1:6]
valsg <- vals[ grep("log.rich", vals$coef),] %>% 
  mutate(term=c("quadrat", "sp", "quadrat:sp"))
test <- summary(mort2)
vals <- as.data.frame(test$coef.mat)[1:6,]
vals$coef <- names(test$coefficients)[1:6]
valsm <- vals[ grep("log.rich", vals$coef),] %>% 
  mutate(term=c("quadrat", "sp", "quadrat:sp"))
test <- summary(rec2)
vals <- as.data.frame(test$coef.mat)[1:6,]
vals$coef <- names(test$coefficients)[1:6]
valsr <- vals[ grep("log.rich", vals$coef),] %>% 
  mutate(term=c("quadrat", "sp", "quadrat:sp")) 

pvals <- bind_rows(list(grow=valsg,mort=valsm,rec=valsr), .id="vital") %>%
  dplyr::select(vital,`Pr(>|z|)`, term) %>% rename(P = `Pr(>|z|)`)
pvals <- rbind(pvals,res) %>% 
  mutate(term = fct_relevel(term, "quadrat", "quadrat:sp", "sp", "Residual"))

pvals$x <- 300
pvals$y <- 0.70
pvals$sig <- paste0("p = ", round(pvals$P,3))
#pvals$sig[pvals$P >0.0166] <- "ns"
pvals$sig[pvals$term=="Residual"] <- ""
pvals$sig[pvals$sig=="p = 0"] <- "p < 0.0001"
```

```{r}
library(wesanderson) #better colour palette
# Gradient color para latitutde
pal <- wes_palette("Zissou1",20, type = "continuous")[20:1] # azul frio-temperado
```


```{r}
fdiri_lat <-ggplot(predis, aes(x=rich.o, y=pred)) +
  geom_line(size=1) +
  facet_grid(vital~term, labeller=grpvit) +
  geom_smooth(data=quants,aes(x=rich.o, y=lower), se=F, size=0.1)+
  geom_smooth(data=quants,aes(x=rich.o, y=upper), se=F, size=0.1)  +
  geom_ribbon(data=quants,aes(x=rich.o, ymin=lower, ymax=upper,
                              y=mean), alpha=0.05, fill="blue",
              size=0)+
  geom_point(data=pontos, aes(x=richness.rarefaction, y=VPC, fill=abs(lat)),
             col="black", size=2.5, pch=21)+
  scale_fill_gradientn(colors=pal) +
  scale_y_continuous(minor_breaks = NULL) +
  scale_x_log10() +
  geom_text(data=pvals, aes(x=x,y=y, label=sig), size=3, hjust=1,vjust=0,
            col="black")+
  
  xlab("Species richness (log10 scale)")+
  ylab("VPC") +
  theme_cowplot() +
  theme(panel.background = element_rect(colour="lightgray"),
        legend.position = "none")
fdiri_lat
png(here("figures/FIG_S5.5_dirichlet_regressions_lat_exclude.png"), height = 600, width=800, res=100)
fdiri_lat
dev.off()
```
