---
title: "RESULTS NOTIME excluding/regrouping rare species"
author: "Melina Leite"
date: "`r  Sys.Date()`"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "..") })
output:
  rmdformats::readthedown:
  highlight: kate
self_contained: true
thumbnails: false
lightbox: true
gallery: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = T, warning = F, fig.align = "center", message = F,
  error = F, cache = F, cache.lazy = FALSE)

library(knitr)
library(tidyverse); library(cowplot); library(patchwork); library(ggrepel)
theme_set(theme_cowplot())
library(ggbeeswarm)
library(DirichletReg)
library(here)
```


Running models with different datasets:

-   **all**: original data, all species all individuals

-   **exclude**: exclude rare species

-   **regroup**: regrouping observations of rare species in ONE species group

Using only the **5x5 m quadrat scale**

```{r run}
#Figures colors/labels
cores <- c("#c00000", "#b666d2", "#70ad47", "#ffc000")

labvit <- as_labeller(c(grow = "Growth", mort="Mortality", 
                        rec="Recruitment"))
grplab = as_labeller(c(quadrat = "Space", `quadrat:sp`= "Space X Species",
                       sp="Species", Residual = "residual"))
grpvit <- as_labeller(c(grow = "Growth", mort="Mortality", 
                        rec="Recruitment",
                        quadrat = "Space", `quadrat:sp`= "Species X Space",
                        sp="Species", Residual = "Residual"))
```

# Data

Growth

```{r}
grp <- c("all","exclude", "regroup")
tudo <- list()
path1 = here("models_outputs/models_notime/")
for (i in 1:length(grp)){
  path <- paste0(path1, grp[i], "/grow/table")
  files = list.files(path)
  all <-  lapply(files, function(x) get(assign(x, load(paste0(path, "/", x)))))
  names(all) <- substr(files,1,nchar(files)-6)
  tudo[[i]] <- all
}
names(tudo) <- grp
tes <- map(tudo,bind_rows, .id="id") %>% bind_rows() %>% filter(q_size=="quad_5")

#tes %>% count(data,fplot) %>% mutate(n=n/8)
table(tes$fplot,tes$data)/5

#remove intercept to include again as a separate column
intercepts <- tes %>% filter(term == "intercept")
names(intercepts)[9] = "intercept"

grow <- tes %>% filter(term != "intercept") %>%
  left_join(intercepts[,c(1:4,9)], by= c("id", "data", "fplot", "time")) %>%
  group_by(id, data, fplot, time) %>% 
  mutate(VPC = variance/sum(variance))  %>% ungroup() %>%
  mutate(term = fct_relevel(term, "quadrat", "quadrat:sp", "sp", "Residual")) %>%
  as.data.frame()

# divide the sd of the terms by the mean growth (intercept) for each dataset.
grow$stdev = grow$Estimate/grow$intercept
```

Mort

```{r}
grp <- c("all","exclude", "regroup")
tudo <- list()
path1 = here("models_outputs/models_notime/")
for (i in 1:length(grp)){
  path <- paste0(path1,grp[i], "/mort/table")
  files = list.files(path)
  all <-  lapply(files, function(x) get(assign(x, load(paste0(path, "/", x)))))
  names(all) <- substr(files,1,nchar(files)-6)
  tudo[[i]] <- all
}
names(tudo) <- grp
tes <- map(tudo,bind_rows, .id="id") %>% bind_rows() %>% filter(q_size=="quad_5")

#tes %>% count(data,fplot) %>% mutate(n=n/8)
table(tes$fplot,tes$data)/5

#remove intercept to include again as a separate column
intercepts <- tes %>% filter(term == "intercept")
names(intercepts)[9] = "intercept"

mort <- tes %>% filter(term != "intercept") %>%
  left_join(intercepts[,c(1:4,9)], by= c("id", "data", "fplot", "time")) %>%
  group_by(id, data, fplot, time) %>% 
  mutate(VPC = variance/sum(variance))  %>% ungroup() %>%
  mutate(term = fct_relevel(term, "quadrat", "quadrat:sp", "sp", "Residual")) %>%
  as.data.frame()
mort$stdev = mort$Estimate
```

Rec

```{r}
grp <- c("all","exclude", "regroup")
tudo <- list()
path1 = here("models_outputs/models_notime/")
for (i in 1:length(grp)){
  path <- paste0(path1,grp[i], "/rec/table")
  files = list.files(path)
  all <-  lapply(files, function(x) get(assign(x, load(paste0(path, "/", x)))))
  names(all) <- substr(files,1,nchar(files)-6)
  tudo[[i]] <- all
}
names(tudo) <- grp
tes <- map(tudo,bind_rows, .id="id") %>% bind_rows() %>% filter(q_size=="quad_5")

#tes %>% count(data,fplot) %>% mutate(n=n/8)
table(tes$fplot,tes$data)/5

#remove intercept to include again as a separate column
intercepts <- tes %>% filter(term == "intercept")
names(intercepts)[9] = "intercept"

rec <- tes %>% filter(term != "intercept") %>%
  left_join(intercepts[,c(1:4,9)], by= c("id", "data", "fplot", "time")) %>%
  group_by(id, data, fplot, time) %>% 
  mutate(VPC = variance/sum(variance))  %>% ungroup() %>%
  mutate(term = fct_relevel(term, "quadrat", "quadrat:sp", "sp", "Residual")) %>%
  as.data.frame()
rec$stdev = rec$Estimate
```

Excluding recruitment for Wytham Woods, as in "all data".
```{r}
rec <- rec %>% filter(fplot!="wyw")
```


Combining results.
```{r}
comb <- bind_rows(list(grow=grow, mort=mort, rec=rec), .id="vital") %>%
  mutate(term = fct_relevel(term, "quadrat", "quadrat:sp", "sp","Residual")) %>%
  ungroup() 
```




# Comparing groups

```{r, fig.height=10, fig.width=20}
comb %>% 
  ggplot(aes(x=data, y=VPC)) +
  geom_col(aes(fill=term, color=term), alpha=0.7) +
  scale_fill_manual(values=cores) +
  scale_color_manual(values=cores) +
  facet_grid(vital~fplot+time) +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

mean over fplots

```{r}
comb %>% group_by(vital,data,term) %>% summarise(VPC=mean(VPC)) %>%
  ggplot(aes(x=data, y=VPC)) +geom_col(aes(fill=term, color=term), alpha=0.7) +
  scale_fill_manual(values=cores) + 
  scale_color_manual(values=cores) +
  facet_grid(~vital)+
  theme(axis.text.x = element_text(angle=45, hjust=1))+
  labs(tag="a)")
#ggsave("figs/FIG_S4.4a_groups_mean.jpg", width=20, height = 10, units = "cm")
```

```{r, fig.height=8}
comb %>%
  ggplot(aes(x=data,y=VPC, col=data)) + geom_boxplot()+
  facet_grid(vital~term) +
  theme(axis.text.x = element_text(angle=45,hjust=1),
        legend.position = "none")
```

```{r}
comb %>% mutate(xis = as.numeric(as.factor(data))) %>%
  ggplot(aes(x=xis,y=stdev, col=paste(fplot,time))) + geom_line() +
  facet_grid(vital~term) + 
  scale_x_continuous(breaks = 1:3, labels=levels(comb$data),
                     name="") +
  theme(axis.text.x = element_text(angle=45, hjust=1),
        legend.position = "none")
```

```{r}
comb %>% mutate(xis = as.numeric(as.factor(data))) %>%
  ggplot(aes(x=xis,y=VPC, col=paste(fplot,time))) + geom_line() +
  facet_grid(vital~term) + 
  scale_x_continuous(breaks = 1:3, labels=levels(comb$data),
                     name="") +
  theme(axis.text.x = element_text(angle=45, hjust=1),
        legend.position = "none")
```



## Difference to all data

### Absolute difference

```{r}
dif <- comb %>% 
  pivot_wider(id_cols = c(vital,fplot,time,term),
              names_from = data,
              values_from = c(stdev,VPC, richness)) %>%
  mutate(stdev.dif_exclude = stdev_exclude - stdev_all,
         VPC.dif_exclude =  VPC_exclude - VPC_all,
         stdev.dif_regroup = stdev_regroup - stdev_all,
         VPC.dif_regroup =  VPC_regroup - VPC_all) %>%
  select(vital,fplot,time,term,richness_exclude,
         stdev.dif_exclude,
         stdev.dif_regroup, VPC.dif_exclude, VPC.dif_regroup) %>%
  pivot_longer(6:9, names_to = c("index", "data"), names_sep = "_",
                #names_pattern = "(.*_).(.*)",
                values_to =  "value") %>%
  pivot_wider(names_from = index, values_from = value)
```

```{r, fig.height=6, fig.width=8}
dif %>%
  ggplot(aes(x=data, y=stdev.dif)) + 
  facet_grid(vital~term, scales= "free_x") +
  geom_boxplot() +
  geom_quasirandom(aes(col=fplot)) +
  ylab("SD no.rare - SD all") + xlab("") +
  geom_hline(yintercept = 0, linetype= "dotted")  +
  ggtitle("Difference in SD") +
  theme(legend.position = "right",
        panel.background = element_rect(colour="lightgray"),
        axis.text.x = element_text(angle=45, hjust=1))
```

```{r, fig.height=6, fig.width=8}
dif %>%
  ggplot(aes(x=data, y=VPC.dif)) + 
  facet_grid(vital~term, scales= "free_x") +
  geom_boxplot() +
  geom_quasirandom(aes(col=fplot)) +
  ylab("VPC no.rare - VPC all") + xlab("") +
  geom_hline(yintercept = 0, linetype= "dotted")  +
  ggtitle("Difference in VPC") +
  theme(legend.position = "right",
        panel.background = element_rect(colour="lightgray"),
        axis.text.x = element_text(angle=45, hjust=1))
```

### Relative difference SD

```{r}
difr <- comb %>% 
  pivot_wider(id_cols = c(vital,fplot,time,term),
              names_from = data,
              values_from = c(stdev,VPC, richness)) %>%
  mutate(stdev.dif_exclude = (stdev_exclude - stdev_all)*100/stdev_all,
         stdev.dif_regroup = (stdev_regroup - stdev_all)*100/stdev_all) %>%
  select(vital,fplot,time,term,richness_exclude,
         stdev.dif_exclude,stdev.dif_regroup) %>%
  pivot_longer(6:7, names_to = "data",
                values_to =  "stdev.dif") %>%
  mutate(data = substr(data,11, nchar(data)))
```

```{r, fig.height=6, fig.width=8}
difr %>%
  ggplot(aes(x=data, y=stdev.dif)) + 
  facet_grid(vital~term, scales= "free_x") +
  geom_boxplot() +
  geom_quasirandom(aes(col=fplot)) +
  ylab("Proportional difference (%)") + xlab("") +
  geom_hline(yintercept = 0, linetype= "dotted")  +
  ggtitle("Relative difference in SD") +
  theme(legend.position = "right",
        panel.background = element_rect(colour="lightgray"),
        axis.text.x = element_text(angle=45, hjust=1))
```

```{r}
difr %>% group_by(vital, term, data) %>% summarise(mean=mean(stdev.dif, na.rm=T)) %>%
  pivot_wider(names_from = term, values_from = mean)
```

#### table

Relative differences in SD

```{r}
difr %>% group_by(vital, term, data) %>% 
  summarise(mean=mean(stdev.dif, na.rm=T)) %>%
  pivot_wider(names_from = c(vital,data), values_from = mean) %>%
  kable(digits=1)
```

Mean absolute differences in VPC

```{r}
dif %>% group_by(vital, term, data) %>% 
  summarise(mean=mean(VPC.dif, na.rm=T)) %>%
  pivot_wider(names_from = c(vital,data), values_from = mean) -> write
kable(write)
```

## Comparing with NUMBER of rare species

```{r, fig.height=6, fig.width=12}
dif %>%
  ggplot(aes(x=richness_exclude, y=stdev.dif, shape=data)) + 
  geom_point(aes(col=fplot)) + 
  geom_smooth(method="lm", se=F, aes(linetype=data)) +
  facet_grid(vital~term) +
  ylab("Difference in SD to original data") +
  xlab("log N of rare species") +
  geom_hline(yintercept = 0, linetype= "dotted")  +
  ggtitle("Difference in SD") +
  scale_x_log10()+
  theme(axis.text.x = element_text(angle=45, hjust=1),
        )
```

```{r, fig.height=6, fig.width=12}
difr %>%
  ggplot(aes(x=richness_exclude, y=stdev.dif, shape=data)) + 
  geom_point(aes(col=fplot)) + 
  geom_smooth(method="lm", se=F, aes(linetype=data)) +
  facet_grid(vital~term) +
  ylab("Relative difference in SD to original data") +
  xlab("log N of rare species") +
  geom_hline(yintercept = 0, linetype= "dotted")  +
  ggtitle("Proportional difference in SD to original data") +
  theme(axis.text.x = element_text(angle=45, hjust=1),
        panel.background = element_rect(fill="white", color="black"))
```

```{r, fig.height=6, fig.width=12}
pm <- dif %>%
  ggplot(aes(x=richness_exclude, y=VPC.dif, shape=data)) + 
  geom_point(aes(col=fplot)) + 
  facet_grid(vital~term) +
   geom_smooth(method="lm", se=F, aes(linetype=data)) +
  ylab(" Absolute difference in VPC to original data") +
   xlab("log N of rare species") +
  geom_hline(yintercept = 0, linetype= "dotted")  +
  theme(axis.text.x = element_text(angle=45, hjust=1),
        panel.background = element_rect(fill="white", color="black"))
pm
```


# Dirichlet regression for exclude results with species richness

Richness by rarefaction at the 6ha plot size (sampling unit 20x20m)

```{r}
load(here("data", "rarefaction.curves.Rdata"))
rich.rare <- rare[rare$sites ==150,2:3]
colnames(rich.rare)[1] <- "richness.rarefaction"
comb <- comb %>% left_join(rich.rare, by="fplot")
```

Latitude

```{r}
load(here("data", "plots_structure.Rdata"))
comb <- comb %>% left_join(plots.structure[,1:2], "fplot")
```

```{r}
#mean for each forest
mcomb <- comb %>% group_by(vital,data, fplot,term, lat) %>%
  summarise(VPC = mean(VPC),
            stdev = mean(stdev),
            richness.rarefaction = mean(richness.rarefaction))
```

## GROWTH

```{r}
predirig <-  mcomb %>% 
  filter(vital == "grow", data == "exclude") %>%
  select(fplot, term, VPC, richness.rarefaction,lat) 
diridatag <- predirig %>%
  pivot_wider(names_from = term, values_from = VPC) %>% ungroup() %>%
  mutate(log.rich = log(richness.rarefaction),
         log.rich.o=log.rich) %>%
  mutate_at(vars( log.rich), scale)

vpcg <- DR_data(diridatag[,c("quadrat", "sp", "quadrat:sp",  "Residual")])
#plot(vpc)
```

```{r, echo=T}
grow2 <- DirichReg(vpcg~log.rich, data=diridatag, model="alternative", base=4)
summary(grow2)
```

```{r}
newdata <- expand.grid(log.rich = seq(min(diridatag$log.rich),
                                      max(diridatag$log.rich), length.out=10))
pred <- predict(grow2, newdata = newdata, se=T)
confint(grow2)
colnames(pred) <- colnames(vpcg)
newdata$log.rich.o <- newdata$log.rich*sd(diridatag$log.rich.o) +
  mean(diridatag$log.rich.o)
newdata$rich.o <- exp(newdata$log.rich.o)
newpredg <- cbind(newdata,pred) %>% pivot_longer(`quadrat`:`Residual`, names_to="term", values_to="pred")

newpredg %>% mutate(term = fct_relevel(term, "quadrat", "quadrat:sp", "sp",
                                       "Residual")) %>%
  ggplot(aes(x=rich.o, y=pred, col=term)) +
  geom_line() + facet_grid(~term) +
  geom_point(data=predirig, aes(x=richness.rarefaction, y=VPC, col=term)) +
  scale_x_log10() +
  ggtitle("Mod log.rich")
```

Residuals

```{r}
resid <- residuals(grow2, type = "standardized")
resid <-  data.frame(resid[,1:4])
resid$log.rich <- diridatag$log.rich
resid <- resid %>% pivot_longer(1:4, names_to="term", values_to="resid")

ggplot(resid, aes(x=log.rich, y=resid)) + geom_point() +
  facet_grid(~term)
```

## MORT

```{r}
predirim <-  mcomb %>% 
  filter(vital == "mort", data == "exclude") %>%
  select(fplot, term, VPC, richness.rarefaction,lat) 
diridatam <- predirim %>%
  pivot_wider(names_from = term, values_from = VPC) %>% ungroup() %>%
  mutate(log.rich = log(richness.rarefaction),
         log.rich.o=log.rich) %>%
  mutate_at(vars( log.rich), scale)

vpcm <- DR_data(diridatam[,c("quadrat", "sp", "quadrat:sp",  "Residual")])
#plot(vpc)
```

```{r, echo=T}
mort2 <- DirichReg(vpcm~log.rich,data=diridatam, model="alternative", base=4)
summary(mort2)
```

```{r}
newdata <- expand.grid(log.rich = seq(min(diridatam$log.rich),
                                      max(diridatam$log.rich), length.out=10))
pred <- predict(mort2, newdata = newdata, se=T)
confint(mort2)
colnames(pred) <- colnames(vpcm)
newdata$log.rich.o <- newdata$log.rich*sd(diridatam$log.rich.o) +
  mean(diridatam$log.rich.o)
newdata$rich.o <- exp(newdata$log.rich.o)
newpredm <- cbind(newdata,pred) %>% pivot_longer(`quadrat`:`Residual`, names_to="term", values_to="pred")

newpredm %>% mutate(term = fct_relevel(term, "quadrat", "quadrat:sp", "sp",
                                       "Residual")) %>%
  ggplot(aes(x=rich.o, y=pred, col=term)) +
  geom_line() + facet_grid(~term) +
  geom_point(data=predirim, aes(x=richness.rarefaction, y=VPC, col=term)) +
  scale_x_log10() +
  ggtitle("Mod log.rich")
```

Residuals
```{r}
resid <- residuals(mort2, type = "standardized")
resid <-  data.frame(resid[,1:4])
resid$log.rich <- diridatam$log.rich
resid <- resid %>% pivot_longer(1:4, names_to="term", values_to="resid")

ggplot(resid, aes(x=log.rich, y=resid)) + geom_point() +
  facet_grid(~term)
```

## REC

```{r}
predirir <-  mcomb %>% 
  filter(vital == "rec", data == "exclude") %>%
  select(fplot, term, VPC, richness.rarefaction,lat) 
diridatar <- predirir %>%
  pivot_wider(names_from = term, values_from = VPC) %>% ungroup() %>%
  mutate(log.rich = log(richness.rarefaction),
         log.rich.o=log.rich) %>%
  mutate_at(vars( log.rich), scale)

vpcr <- DR_data(diridatar[,c("quadrat", "sp", "quadrat:sp",  "Residual")])
#plot(vpc)
```

```{r}
rec2 <- DirichReg(vpcr~log.rich,data=diridatar, model="alternative", base=4)
summary(rec2)
```

```{r}
newdata <- expand.grid(log.rich = seq(min(diridatar$log.rich),
                                      max(diridatar$log.rich), length.out=10))
pred <- predict(rec2, newdata = newdata, se=T)
colnames(pred) <- colnames(vpcr)
newdata$log.rich.o <- newdata$log.rich*sd(diridatar$log.rich.o) +
  mean(diridatar$log.rich.o)
newdata$rich.o <- exp(newdata$log.rich.o)
newpredr <- cbind(newdata,pred) %>% pivot_longer(`quadrat`:`Residual`, names_to="term", values_to="pred")

newpredr %>% mutate(term = fct_relevel(term, "quadrat", "quadrat:sp", "sp",
                                       "Residual")) %>%
  ggplot(aes(x=rich.o, y=pred, col=term)) +
  geom_line() + facet_grid(~term) +
  geom_point(data=predirir, aes(x=richness.rarefaction, y=VPC, col=term)) +
  scale_x_log10() +
  ggtitle("Mod log.rich")
```

Residuals

```{r}
resid <- residuals(rec2, type = "standardized")
resid <-  data.frame(resid[,1:4])
resid$log.rich <- diridatar$log.rich
resid <- resid %>% pivot_longer(1:4, names_to="term", values_to="resid")

ggplot(resid, aes(x=log.rich, y=resid)) + geom_point() +
  facet_grid(~term)
```

## FIGURE

Calculated prediction interval

```{r}
source(here("scripts/prediction_intervals_dirichlet_exclude.R"), local=T)
#load("results/prediction_intervals_dirichlet_exclude.Rdata") #  quants
```

```{r}
predis <- bind_rows(list(grow=newpredg, mort=newpredm, rec=newpredr),
                    .id="vital") %>% 
  mutate(term = fct_relevel(term, "quadrat", "quadrat:sp", "sp","Residual"))

pontos <- bind_rows(list(grow=predirig, mort= predirim, rec=predirir), .id="vital")%>% 
  mutate(term = fct_relevel(term, "quadrat", "quadrat:sp", "sp", "Residual"))
quants <- bind_rows(list(grow=quantg,mort=quantm, rec=quantr), .id="vital")%>% 
  mutate(term = fct_relevel(term, "quadrat", "quadrat:sp", "sp", "Residual"))
```

```{r}
# pvalues
res <- data.frame(vital = c("grow", "mort", "rec"),
                  P = NA,
                  term = "Residual")

test <- summary(grow2)
vals <- as.data.frame(test$coef.mat)[1:6,]
vals$coef <- names(test$coefficients)[1:6]
valsg <- vals[ grep("log.rich", vals$coef),] %>% 
  mutate(term=c("quadrat", "sp", "quadrat:sp"))
test <- summary(mort2)
vals <- as.data.frame(test$coef.mat)[1:6,]
vals$coef <- names(test$coefficients)[1:6]
valsm <- vals[ grep("log.rich", vals$coef),] %>% 
  mutate(term=c("quadrat", "sp", "quadrat:sp"))
test <- summary(rec2)
vals <- as.data.frame(test$coef.mat)[1:6,]
vals$coef <- names(test$coefficients)[1:6]
valsr <- vals[ grep("log.rich", vals$coef),] %>% 
  mutate(term=c("quadrat", "sp", "quadrat:sp")) 

pvals <- bind_rows(list(grow=valsg,mort=valsm,rec=valsr), .id="vital") %>%
  dplyr::select(vital,`Pr(>|z|)`, term) %>% rename(P = `Pr(>|z|)`)
pvals <- rbind(pvals,res) %>% 
  mutate(term = fct_relevel(term, "quadrat", "quadrat:sp", "sp", "Residual"))

pvals$x <- 300
pvals$y <- 0.70
pvals$sig <- paste0("p = ", round(pvals$P,3))
#pvals$sig[pvals$P >0.0166] <- "ns"
pvals$sig[pvals$term=="Residual"] <- ""
pvals$sig[pvals$sig=="p = 0"] <- "p < 0.0001"
```

```{r}
library(wesanderson) #better colour palette
# Gradient color para latitutde
pal <- wes_palette("Zissou1",20, type = "continuous")[20:1] # azul frio-temperado
```


```{r}
fdiri_lat <-ggplot(predis, aes(x=rich.o, y=pred)) +
  geom_line(size=1) +
  facet_grid(vital~term, labeller=grpvit) +
  geom_smooth(data=quants,aes(x=rich.o, y=lower), se=F, size=0.1)+
  geom_smooth(data=quants,aes(x=rich.o, y=upper), se=F, size=0.1)  +
  geom_ribbon(data=quants,aes(x=rich.o, ymin=lower, ymax=upper,
                              y=mean), alpha=0.05, fill="blue",
              size=0)+
  geom_point(data=pontos, aes(x=richness.rarefaction, y=VPC, fill=abs(lat)),
             col="black", size=2.5, pch=21)+
  scale_fill_gradientn(colors=pal) +
  scale_y_continuous(minor_breaks = NULL) +
  scale_x_log10() +
  geom_text(data=pvals, aes(x=x,y=y, label=sig), size=3, hjust=1,vjust=0,
            col="black")+
  
  xlab("Species richness (log10 scale)")+
  ylab("VPC") +
  theme_cowplot() +
  theme(panel.background = element_rect(colour="lightgray"),
        legend.position = "none")
fdiri_lat
png(here("figures/FIG_S5.5_dirichlet_regressions_lat_exclude.png"), height = 600, width=800, res=100)
fdiri_lat
dev.off()
```
