---
title: "Forest plot data preparation workflow for Barro Colorado Islands"
author: "Melina Leite"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "..") })
output:
  rmdformats::readthedown:
    highlight: kate
    self_contained: true
    thumbnails: false
    lightbox: true
    gallery: false
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  echo = T, warning = F, fig.align = "center", message = F,
  error = F, cache = T, cache.lazy = FALSE)

library(knitr)
library(tidyverse); library(cowplot); library(patchwork); 
theme_set(theme_cowplot())
library(here)
library(janitor)
library(forcats)


```

This script is part of the repository of analysis from **Leite et al. 2022. Major axe of variation in tree demography across global forests**.

In this section, we will work with Barro Colorado Island (BCI) data as an example of the **data cleaning/wrangling pipeline** used in the analysis of the 21 ForestGEO plots.

The source of the data is avalilable at the Dryad repository:

> Condit, Richard et al. (2019), Complete data from the Barro Colorado 50-ha plot: 423617 trees, 35 years, Dryad, Dataset, https://doi.org/10.15146/5xcp-0d46.

As stated in the Abstract of the dataset:

> "The 50-ha plot at Barro Colorado Island was initially demarcated and fully censused in 1982, and has been fully censused 7 times since, every 5 years from 1985 through 2015. Every measurement of every stem over 8 censuses is included in this archive.(..) This is the 2019 version of the database, frozen as of 30 May 2019."


Auxiliary functions for data checking
```{r}
source(here("workflow_example","scripts","functions_checking.R"))
source(here("workflow_example","scripts","functions_status.R"))
```


# 1. Dataset

We will use only the tree level data and the species table.

The data is already in **.Rdata** format.

Plot dimensions.
```{r}
plotdimension = matrix(c(0, 1000, 0, 500), ncol = 2, byrow = T)
plot = list(plotdimension = plotdimension)
plot
```

Species table.
```{r}
load(here("workflow_example", "bci_rawdata","bci.spptable.rdata"))

species = bci.spptable

# Identifying ferns and palms
fernpalms <- c("Arecaceae", "Cyatheaceae", "Dicksoniaceae", "Metaxyaceae",
               "Cibotiaceae","Thyrsopteridaceae", "Loxomataceae", 
               "Culcitaceae", "Plagiogyriaceae")

species$fern.palm <- "FALSE" 
species$fern.palm[species$Family %in% fernpalms] <- "TRUE"

#unidentified species
unind <- species[species$Genus == "Unidentified",]

#morphospecies
morpho <- species[species$IDLevel == "genus",] 
```


Tree data.  
A list with 8 tables one per census.   
Each row is a tree the respective census.  

Columns  

- treeID: identification of the tree  
- tag ( _not used_ ): tag number of the tree in the field  
- sp: species identification code, which matches with species table  
- quadrat ( _not used_ ): quadrat code in the plot  
- gx and gy: coordinates of the tree in the plot  
- stemID ( _not used_ ): stem identification (but the data is at tree level)  
- dbh: diameter at breast height  
- hom: height of measurement of dbh. usualy is 1.3m  
- ExactDate: date of measurement  
- codes ( _not used_ ): codes for tree conditions  
- date ( _not used_ ):
- status: status of the tree (see below the code for status)  
- nostems: number of stems  
- agb ( _not used_ ): aboveground biomass  
- ba: basal area  

Status code:  

- A: alive  
- D: dead  
- M: missing  
- P: prior (to recruitment)

 
Import data to list   
```{r}
path <- here("workflow_example", "bci_rawdata", "bci.tree")
files = list.files(path)
tree = lapply(files, function(x) get(assign(x, load(paste0(path, "/", x)))))
names(tree) = paste0("tree", 1:length(files))
```

Checking data
```{r}
names(tree$tree1)
str(tree$tree8)
```

Save ExactDate as.Date as column date
```{r}
tree = lapply(tree, transform, date = as.Date(ExactDate))
```

Add column to indicate changes in high of measurement (HOM) of DBH (diameter at breast height) from one census to the next.
```{r}
tree[[1]]$homchange = NA
for (i in 2:length(tree)) {
  tree[[i]]$homchange = tree[[i]]$hom != tree[[i-1]]$hom
}
```

Create theoretical dbh based on basal area (ba) for multistem trees
```{r}
tree <- lapply(tree, function(x){
  x$dbh <- as.numeric(2000*sqrt(x$ba/pi))
  x$dbh[which(x$dbh==0)] <- NA
  x$ba <- as.numeric(x$ba)
  x
})
```

Keeping only the used columns.
```{r} 
tree <- lapply(tree, select, c("treeID","sp","gx","gy","dbh","homchange",
                               'ba', "nostems","status","date"))
```

Excluding trees with `NA` in gx and gy coordinates.
```{r} 
tree <- lapply(tree, function(x){
  x %>% filter(!is.na(gx), !is.na(gy))
})
```

Checking column type.
```{r} 
# column type
tree <- lapply(tree, function(x){
  x %>% mutate(treeID=as.character(treeID))
})
```

Checking and correcting status.
```{r} 
tree.orig <- tree
tree <- correct_status(tree)
check_status(tree, tree.orig)
```

Quality checks for tree data
```{r} 
check_dim(tree)
check_col(tree)
check_types(tree)
check_order(tree)
check_units(tree)
check_nas(tree)
```

# 2. Growth data

- Selecting only the individuals that were alive (status =="A") in the initial and final census.

- Removing NAs in species and date, removing ferns and palms.

- Removing individuals with <10mm DHB in the initial census.

- Calculating growth: `g.dbh`

- Calculating time interval between measurements in days. Removing intervals with zero and negative days

- Calculating growth rate / year: growth*365/days

```{r}
n.census <- length(tree)
growth <- list()
for (i in 1:(n.census-1)){
  first <- tree[[i]]
  second <- tree[[i+1]] %>% dplyr::select(-sp, -gx, -gy)
  df <- inner_join(first%>%select(-homchange), second, by=c("treeID"), 
                   suffix= c(".i", ".f"))
  df <- df %>% filter(status.i == "A" & 
                      status.f  == "A") %>%
              filter(dbh.i >=10)
  df$interval <- df$date.f - df$date.i
  df$g.ba <- (df$ba.f-df$ba.i)*365/as.numeric(df$interval)
  df$g.dbh <- (df$dbh.f-df$dbh.i)*365/as.numeric(df$interval)
  df <- df %>% filter(!is.na(g.dbh), !is.na(g.ba), !is.na(sp), !is.na(interval),
                      interval>0, sp %in% species$sp[species$fern.palm == "FALSE"])
  growth[[i]] <- df
}
```

Removing trees with different HOM.
```{r}
growth <- lapply(growth, function(x) {
 x[which(x$homchange==FALSE), ]
})
```


Including Quadrat scales columns.
```{r}
scales <- c(5,10,20,50,100)

for (g in 1:(n.census-1)) {
  dad <- growth[[g]]
  
  escas <- as.data.frame(matrix(ncol=length(scales), nrow=nrow(dad)))
  
   for (k in 1:length(scales)){
    scala <- scales[k]
    xis <- c(0, cumsum(rep(scala, plot[[1]][1,2]/scala)))
    yis <- c(0, cumsum(rep(scala, plot[[1]][2,2]/scala)))
    
    qxis <- qyis <- rep(NA, length=nrow(dad))
    
    for (i in 1:(length(xis)-1)){
      for (n in 1:(length(yis)-1)){
        qxis[dad$gx >= xis[i] & dad$gx <= xis[i+1]] <- xis[i]
        qyis[dad$gy >= yis[n] & dad$gy <= yis[n+1]] <- yis[n] 
      }
    }
    colnames(escas)[k] <- paste("quad", scala, sep="_")
    escas[,k] <- paste(qxis, qyis, sep="_")
    escas[grep("NA", escas[,k]),k] <- NA
   }
  growth[[g]] <- cbind(dad, escas)
}
```


Removing the extreme trees with more than 4 standard deviations of growth for both tails. It decided after some tests with simulated measurement error data and following literature reccommendations, as in Visser et al. 2016:

>"For data set 1, we excluded individuals marked as ‘resprout’, ‘buttressed’, ‘leaning’ and ‘broken above 1.3 m’ in each census, as well as those with growth rates more than four 4 standard deviations from the mean. These are likely measurement errors (Ru€ger et al. 2011)."

```{r}
dbh.4sd <- lapply(growth, function(tes){
  tes[tes$g.dbh < 4*sd(tes$g.dbh) & tes$g.dbh > -4*sd(tes$g.dbh), ]
})
```

Resuming table of number of individuals and species in each census interval (time).

```{r}
ndbh.4sd <- map_df(dbh.4sd,  function(x){
   data.frame(Nsp = length(unique(x$sp)), N = nrow(x))
}, .id="time")

ndata <-  ndbh.4sd %>% group_by(time) %>%
  mutate(perc.loss.Nsp = 1-round(Nsp/Nsp[1],2),
         perc.loss.N = 1-round(N/N[1],2)) %>%
  select(time,Nsp, perc.loss.Nsp, N, perc.loss.N ) %>%
  arrange(time)
kable(ndata)
```

Saving growth data object.
```{r}
file <- paste0(here("workflow_example", "bci_cleandata"),"/growth.rds")
saveRDS(dbh.4sd, file=file)
```

# 3. Mortality data

- Remove trees with NA in species names, ferns and palms; NA, zero or negative date interval 

- Selecting only the individuals that were alive (status "A") in the initial census and Alive or Dead (status "A" or "D") in the final census.

- Create a column to identify if the tree is alive (0) or dead (1)

```{r}
mort <- list()
for (i in 1:(n.census-1)){
  first <- tree[[i]]
  second <- tree[[i+1]] %>% dplyr::select(-sp, -gx, -gy)
  df <- inner_join(first, second, by=c("treeID"), 
                   suffix= c(".i", ".f"))
  df <- df %>% filter(status.i == "A" & 
                      status.f  %in% c("A", "D"))%>%
              filter(dbh.i >=10)

  df$interval <- df$date.f - df$date.i
  df$dead <- ifelse(df$status.f == 'D', 1, 0)
  df <- df %>% filter(!is.na(sp), !is.na(interval), interval>0,
                      sp %in% species$sp[species$fern.palm == "FALSE"])
  mort[[i]] <- df
}
```

Including Quadrat scales columns.
```{r}
scales <- c(5,10,20,50,100)
for (g in 1:(n.census-1)) {
  dad <- mort[[g]]
  
  escas <- as.data.frame(matrix(ncol=length(scales), nrow=nrow(dad)))
  
   for (k in 1:length(scales)){
    scala <- scales[k]
    xis <- c(0, cumsum(rep(scala, plot[[1]][1,2]/scala)))
    yis <- c(0, cumsum(rep(scala, plot[[1]][2,2]/scala)))
    
    qxis <- qyis <- rep(NA, length=nrow(dad))
    
    for (i in 1:(length(xis)-1)){
      for (n in 1:(length(yis)-1)){
        qxis[dad$gx >= xis[i] & dad$gx <= xis[i+1]] <- xis[i]
        qyis[dad$gy >= yis[n] & dad$gy <= yis[n+1]] <- yis[n] 
      }
    }
    colnames(escas)[k] <- paste("quad", scala, sep="_")
    escas[,k] <- paste(qxis, qyis, sep="_")
    escas[grep("NA", escas[,k]),k] <- NA
   }
  mort[[g]] <- cbind(dad, escas)
}
```


Summary mortality tables.

**Mortality rate is $m_a$ sensu Kohyama et al. 2018**

$$m_a = 1 - \left(\frac{N_{ST}}{N_0}\right)^\frac{1}{T}$$

$N_0$ number of alive individuals initial census  
$N_ST$ individuals that survived from the intial to the final census  
$T$ time interval in years  

```{r}
map_df(mort,  function(x){
  y <-  data.frame(Nsp = length(unique(x$sp)), 
              N_0 = nrow(x),
              N_dead = sum(x$dead), 
              N_ST = nrow(x)-sum(x$dead),
              mean_T = as.numeric(mean(x$interval, na.rm=T))/365)
  y$mort.rate <- 1 - (y$N_ST/y$N_0)^(1/y$mean_T)
  return(y)
}, .id="time") %>% kable()
```

**OBS**: mean time interval in years between census **for all individuals**


Saving mort data object.
```{r}
file <- paste0(here("workflow_example", "bci_cleandata"), "/mortality.rds")
saveRDS(mort, file=file)
```

# 4. Recruitment 

1. Remove trees with NA in species names, ferns and palms, and with NA in the final date.

2. - Selecting only the individuals that were alive in final census (status "A") and alive or recruits in the first census ("A", "P").

3. Removing individuals with <10mm DHB in the first census for the alive individuals and in the second census for the priors only for the priors (final census).

4. Create a column to identify if the tree is alive (0) or recruit (1)

```{r}
rec <- list()
for (i in 1:(n.census-1)){
  first <- tree[[i]]
  second <- tree[[i+1]] %>% dplyr::select(-sp, -gx, -gy)
  df <- inner_join(first, second, by=c("treeID"), 
                   suffix= c(".i", ".f"))
  df <- df %>% filter(status.f =="A",
                      status.i %in% c("A","P")) 
  df <- df %>% filter(dbh.i >=10 | is.na(dbh.i)) %>%
               filter(status.i != "P" | dbh.f >=10)

  df$interval <- df$date.f - df$date.i
  df$rec <- ifelse(df$status.i == 'P', 1, 0)
  
  # ferns & palms
  df <- df %>% filter(!is.na(sp), sp %in% species$sp[species$fern.palm == "FALSE"])
  
  df <- df %>% filter(!is.na(date.f)) # excluding NA in final date
  
  rec[[i]] <- df
}
```

Including Quadrat scales columns.
```{r}
scales <- c(5,10,20,50,100)
for (g in 1:(n.census-1)) {
  dad <- rec[[g]]
  
  escas <- as.data.frame(matrix(ncol=length(scales), nrow=nrow(dad)))
  
   for (k in 1:length(scales)){
    scala <- scales[k]
    xis <- c(0, cumsum(rep(scala, plot[[1]][1,2]/scala)))
    yis <- c(0, cumsum(rep(scala, plot[[1]][2,2]/scala)))
    
    qxis <- qyis <- rep(NA, length=nrow(dad))
    
    for (i in 1:(length(xis)-1)){
      for (n in 1:(length(yis)-1)){
        qxis[dad$gx >= xis[i] & dad$gx <= xis[i+1]] <- xis[i]
        qyis[dad$gy >= yis[n] & dad$gy <= yis[n+1]] <- yis[n] 
      }
    }
    colnames(escas)[k] <- paste("quad", scala, sep="_")
    escas[,k] <- paste(qxis, qyis, sep="_")
    escas[grep("NA", escas[,k]),k] <- NA
   }
  rec[[g]] <- cbind(dad, escas)
}
```


Summary recruitment tables

**recruitment rate is $r_af$ sensu Kohyama et al. 2018:**

$$r_{af} = 1 - \left(\frac{N_{ST}}{N_T}\right)^\frac{1}{T}$$

$N_T$ total of alive individuals in the final census  
$N_ST$ individuals that survived from the intial to the final census  
$T$ time interval in years  

```{r}
map_df(rec,  function(x){
  y <-  data.frame(Nsp = length(unique(x$sp)), 
              N_T = nrow(x),
              N_recruits = sum(x$rec), 
              N_ST = nrow(x)-sum(x$rec),
              mean_T = as.numeric(mean(x$interval, na.rm=T))/365)
  y$rec.rate <- 1 - (y$N_ST/y$N_T)^(1/y$mean_T)
  return(y)
}, .id="time") %>% kable()
```

**OBS**: mean time interval in years between census **for the individuals that survived**

Saving recruitment data object.
```{r}
file <- paste0(here("workflow_example", "bci_cleandata"), "/recruitment.rds")
saveRDS(rec, file=file)
```
