---
title: "Modeling workflow for Barro Colorado Islands"
subtitle: "Analysis from Leite et al. 2022"
author: "Melina Leite"
date: "`r Sys.Date()"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "..") })
output:
  rmdformats::readthedown:
    highlight: kate
    self_contained: true
    thumbnails: false
    lightbox: true
    gallery: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = T, warning = F, fig.align = "center", message = F,
  error = F, cache = F, cache.lazy = FALSE)

library(knitr)
library(tidyverse); library(cowplot); library(patchwork); 
theme_set(theme_cowplot())
library(here)
library(future.apply)
library(brms)
library(ggmcmc)
```


This script is part of the repository of analysis from **Leite et al. 2022. Major axe of variation in tree demography across global forests**.

In this section, we will work with Barro Colorado Island (BCI) data as an example of the **modeling pipeline** used in the analysis of the 21 ForestGEO plots.

For data cleaning/wrangling, please see the previous script `1_dataPreparation.Rmd`.

For the example analyses, we used the 100 x 100 m (1ha) quadrat grain size, which is the fastest scale for models to run.

# No-time/reduced models

For the example of models without temporal organizing principles (reduced models), we use the last cenus interval (7).

## 1. Growth reduced model

Loading the cleaned data.
```{r}
df <- readRDS(here("workflow_example", "bci_cleandata", "growth.rds"))
dad <- bind_rows(df, .id = "time")
data <- dad[dad$time == 7, ]
data$quadrat <- data[,colnames(data) == "quad_100"]
```

Models parameters.
```{r}
delta = 0.95
ncores = 3 # number of chains
iter = 3000 # number of iterations per chain
warmup = 1000 # number of burning iterations discarded per chain
thin = 5 # thinning interval
```

Model.
```{r, eval=F}
gm <- brm(g.dbh ~ 1 + (1|quadrat) + (1|sp) + (1|quadrat:sp), 
              data=data, 
              chains = ncores,
              control = list(adapt_delta = delta),
              cores = ncores,
              iter = iter, warmup=warmup,
              thin = thin, 
              seed=T)
save(gm, file = here("workflow_example", "bci_models_outputs","no_time_models",
          "grow", "bci-7-quad_100-model.Rdata"))
```

It took more than 9 hours to run the model in a high performance cluster with parallel chains. I do not recommend you to run it in your personal computer. The model is already saved in the `bci_models_output` folder as an .Rdata file (>320MB):
```{r load grow}
load(here("workflow_example", "bci_models_outputs","no_time_models",
          "grow", "bci-7-quad_100-model.Rdata"))
```

**Warning**: If you cloned this repository originally from [GitHub](https://github.com/melina-leite/varpart_treeVitalRates), and the chunk above doesn't work, it is probably because you need to instal [git large files storage (LFS)](https://git-lfs.github.com/) before cloning it in order to download the model's files correctly.


Summary table with posterior medians.
```{r}
gmsum <- summary(gm, robust=T) # median posterior
gmsum
```

Traceplots for the posterior distribution of parameters and chains mixture evaluation. Also saved as pdf A4 files.
```{r trace grow}
pdf(file=here("workflow_example", "bci_models_outputs","no_time_models",
          "grow", "bci-7-quad_100-traceplot.pdf"), 
    paper="a4", height = 20, width=10, onefile=F)
plot(gm, ask=F, N=5)
dev.off()
```


```{r}
plot(gm, ask=F, N=5)
```

Table of model results.
```{r}
fix <- as.data.frame(gmsum$fixed)
res <- as.data.frame(gmsum$spec_pars)
quadrat <- as.data.frame(gmsum$random$quadrat)
sp <- as.data.frame(gmsum$random$sp)
quadrat.sp <- as.data.frame(gmsum$random$`quadrat:sp`)

result <- bind_rows(list(intercept = fix, 
             quadrat = quadrat,
             sp = sp,
             `quadrat:sp` = quadrat.sp,
             Residual = res), .id="term")
rownames(result) <- NULL

result$levels = c(NA, gmsum$ngrps$quadrat[1],
         gmsum$ngrps$sp[1],
         gmsum$ngrps$`quadrat:sp`[1],
         gmsum$nobs)
result$variance <- result$Estimate^2

sdtab <- data.frame(data = "all", 
                  fplot = "bci",
                  time = "7",
                  q_size = "quad_100",
                  ntrees = gmsum$nobs,
                  richness = gmsum$ngrps$sp[1])
sdtab <- sdtab[rep(seq_len(nrow(sdtab)), each = 5), ]
rownames(sdtab) <- NULL

res <- cbind(sdtab, result)
kable(res, digits = 2)
# saving table ------------
save(res, file=here("workflow_example", "bci_models_outputs","no_time_models",
          "grow", "bci-7-quad_100-table.Rdata"))
```

## 2. Mortality reduced model

Loading the cleaned data.
```{r}
df <- readRDS(here("workflow_example", "bci_cleandata", "mortality.rds"))
dad <- bind_rows(df, .id = "time")
data <- dad[dad$time == 7, ]
data$quadrat <- data[,colnames(data) == "quad_100"]
```

Models parameters.
```{r}
delta=0.95
ncores = 3 # number of chains
iter = 3000 # number of iterations per chain
warmup = 1000 # number of burning iterations discarded per chain
thin = 5 # thinning interval
```

Model.
```{r, eval=F}
gm <- brm(dead ~ 1 + (1|quadrat) + (1|sp) + (1|quadrat:sp) +
                offset(log(y.interval)),
              family = bernoulli(link="cloglog"),
              data = data, 
              chains = ncores,
              control = list(adapt_delta = delta),
              cores = ncores,
	            iter = iter, warmup=warmup,
              thin = thin)
save(gm, file = here("workflow_example", "bci_models_outputs","no_time_models",
          "mort", "bci-7-quad_100-model.Rdata"))
```

The model is already saved in the `bci_models_output` folder.
```{r}
load(here("workflow_example", "bci_models_outputs", "no_time_models",
          "mort", "bci-7-quad_100-model.Rdata"))
```

Summary table with posterior medians.
```{r}
gmsum <- summary(gm, robust=T) # median posterior
gmsum
```

Traceplots for the posterior distribution of parameters and chains mixture evaluation. Also saved as pdf A4 files.
```{r trace mort}
pdf(file=here("workflow_example", "bci_models_outputs","no_time_models",
          "mort", "bci-7-quad_100-traceplot.pdf"), 
    paper="a4", height = 20, width=10, onefile=F)
plot(gm, N=4, ask=F)
dev.off()
```

```{r}
plot(gm, ask=F, N=4)
```

Results table.
```{r}
fix <- as.data.frame(gmsum$fixed)
res <- as.data.frame(gmsum$spec_pars)
res[1,] <- c(sqrt((pi^2)/6), rep(NA,6))
quadrat <- as.data.frame(gmsum$random$quadrat)
sp <- as.data.frame(gmsum$random$sp)
quadrat.sp <- as.data.frame(gmsum$random$`quadrat:sp`)

result <- bind_rows(list(intercept = fix, 
             quadrat = quadrat,
             sp = sp,
             `quadrat:sp` = quadrat.sp,
             Residual = res), .id="term")
rownames(result) <- NULL

result$levels = c(NA, gmsum$ngrps$quadrat[1],
         gmsum$ngrps$sp[1],
         gmsum$ngrps$`quadrat:sp`[1],
         gmsum$nobs)
result$variance <- result$Estimate^2

sdtab <- data.frame(data = "all", 
                  fplot = "bci",
                  time = "7",
                  q_size = "quad_100",
                  ntrees = gmsum$nobs,
                  richness = gmsum$ngrps$sp[1])
sdtab <- sdtab[rep(seq_len(nrow(sdtab)), each = 5), ]
rownames(sdtab) <- NULL

res <- cbind(sdtab, result)
kable(res, digits = 2)
# saving table ------------
save(res, file=here("workflow_example", "bci_models_outputs","no_time_models",
          "mort", "bci-7-quad_100-table.Rdata"))
```


## 3. Recruitment reduced model

Loading the cleaned data.
```{r}
df <- readRDS(here("workflow_example", "bci_cleandata", "recruitment.rds"))
dad <- bind_rows(df, .id = "time")
data <- dad[dad$time == 7, ]
data$quadrat <- data[,colnames(data) == "quad_100"]
```

Models parameters.
```{r}
delta=0.95
ncores = 3 # number of chains
iter = 3000 # number of iterations per chain
warmup = 1000 # number of burning iterations discarded per chain
thin = 5 # thinning interval
```

Model.
```{r, eval=F}
gm <- brm(dead ~ 1 + (1|quadrat) + (1|sp) + (1|quadrat:sp) +
                offset(log(y.interval)),
              family = bernoulli(link="cloglog"),
              data = data, 
              chains = ncores,
              control = list(adapt_delta = delta),
              cores = ncores,
	            iter = iter, warmup=warmup,
              thin = thin)
save(gm, file = here("workflow_example", "bci_models_outputs","no_time_models",
          "rec", "bci-7-quad_100-model.Rdata"))
```

The model is already saved in the `bci_models_output` folder.
```{r}
load(here("workflow_example", "bci_models_outputs","no_time_models",
          "rec", "bci-7-quad_100-model.Rdata"))
```

Summary table with posterior medians.
```{r}
gmsum <- summary(gm, robust=T) # median posterior
gmsum
```

Traceplots for the posterior distribution of parameters and chains mixture evaluation. Also saved as pdf A4 files.
```{r}
pdf(file=here("workflow_example", "bci_models_outputs","no_time_models",
          "rec", "bci-7-quad_100-traceplot.pdf"), 
    paper="a4", height = 20, width=10, onefile=FALSE)
plot(gm)
dev.off()
```

```{r}
plot(gm, ask=F, N=4)
```

Table of model results.
```{r}
fix <- as.data.frame(gmsum$fixed)
res <- as.data.frame(gmsum$spec_pars)
res[1,] <- c(sqrt((pi^2)/6), rep(NA,6))
quadrat <- as.data.frame(gmsum$random$quadrat)
sp <- as.data.frame(gmsum$random$sp)
quadrat.sp <- as.data.frame(gmsum$random$`quadrat:sp`)

result <- bind_rows(list(intercept = fix, 
             quadrat = quadrat,
             sp = sp,
             `quadrat:sp` = quadrat.sp,
             Residual = res), .id="term")
rownames(result) <- NULL

result$levels = c(NA, gmsum$ngrps$quadrat[1],
         gmsum$ngrps$sp[1],
         gmsum$ngrps$`quadrat:sp`[1],
         gmsum$nobs)
result$variance <- result$Estimate^2

sdtab <- data.frame(data = "all", 
                  fplot =  "bci",
                  time = "7",
                  q_size = "quad_100",
                  ntrees = gmsum$nobs,
                  richness = gmsum$ngrps$sp[1])
sdtab <- sdtab[rep(seq_len(nrow(sdtab)), each = 5), ]
rownames(sdtab) <- NULL

res <- cbind(sdtab, result)
kable(res, digits = 2)
# saving table ------------
save(res, file=here("workflow_example", "bci_models_outputs","no_time_models",
          "rec", "bci-7-quad_100-table.Rdata"))
```

# Time/complete models

Here, we show as example only one of the 10 subsampled data models.

Loading the code for the first random 5-1ha quadrats:
```{r}
load(here("data", "samples_5ha.Rdata"))
samples$bci$bci_1 # quadrats
```


## 1. Growth time model

Loading the cleaned data.
```{r}
df <- readRDS(here("workflow_example", "bci_cleandata", "growth.rds"))
data <- bind_rows(df, .id = "time")
data$quadrat <- data[,colnames(data) == "quad_100"]
```

For growth in BCI, we excluded the first census interval given measurement problems in the first census.
```{r}
data <- data %>% filter(time != "1")
```

Taking the subsample.
```{r}
data <- data[data$quad_100 %in% samples$bci$bci_1, ]
```

Models parameters
```{r}
delta=0.95
ncores = 3 # number of chains
iter = 3000 # number of iterations per chain
warmup = 1000 # number of burning iterations discarded per chain
thin = 5 # thinning interval
```

Model.
```{r, eval=F}
gm <- brm(g.dbh ~ 1 + (1|quadrat) + (1|sp) + (1|time) +
              (1|quadrat:sp) + (1|quadrat:time) + (1|sp:time), 
            data=data, 
            chains = ncores,
            control = list(adapt_delta = delta),
            cores = ncores,
            iter = iter, warmup=warmup,
            thin = thin,
            save_pars = save_pars(group=FALSE),
            seed=T)
save(gm, file = here("workflow_example", "bci_models_outputs","time_models",
          "grow", "bci-quad_100-1-model.Rdata"))
```

This model (as all the time models) takes more than 4 days to run in a high performance cluster. I don't advise you to run it in your personal computer. The model is already saved in the `bci_models_output` folder.
```{r}
load(here("workflow_example", "bci_models_outputs","time_models",
          "grow", "bci-quad_100-1-model.Rdata"))
```

Summary table with posterior medians.
```{r}
gmsum <- summary(gm, robust=T) # median posterior
gmsum
```

Traceplots for the posterior distribution of parameters and chains mixture evaluation. Also saved as pdf A4 files.
```{r, eval=F}
pdf(file=here("workflow_example", "bci_models_outputs","time_models",
          "grow", "bci-quad_100-1-traceplot.pdf"), 
    paper="a4", height = 20, width=10,
    onefile = T)
plot(gm, ask=F, N=4)
dev.off()
```

```{r}
plot(gm, ask=F, N=4)
```

Table of model results.
```{r}
fix <- as.data.frame(gmsum$fixed)
  res <- as.data.frame(gmsum$spec_pars)
  
  quadrat <- as.data.frame(gmsum$random$quadrat)
  sp <- as.data.frame(gmsum$random$sp)
  time <- as.data.frame(gmsum$random$time)
  quadrat.sp <- as.data.frame(gmsum$random$`quadrat:sp`)
  quadrat.time <- as.data.frame(gmsum$random$`quadrat:time`)
  sp.time <- as.data.frame(gmsum$random$`sp:time`)
  
  result <- bind_rows(list(intercept = fix, 
                           quadrat = quadrat,
                           sp = sp,
                           time = time,
                           `quadrat:sp` = quadrat.sp,
                           `quadrat:time` = quadrat.time,
                           `sp:time` = sp.time,
                           Residual = res), .id="term")
  rownames(result) <- NULL
  
  
  result$levels = c(NA, gmsum$ngrps$quadrat[1],
                    gmsum$ngrps$sp[1],
                    gmsum$ngrps$time[1],
                    gmsum$ngrps$`quadrat:sp`[1],
                    gmsum$ngrps$`quadrat:time`[1],
                    gmsum$ngrps$`sp:time`[1],
                    gmsum$nobs)
  result$variance <- result$Estimate^2
  
  sdtab <- data.frame(sub = "1",
                      data = "all", 
                      fplot = "bci",
                      q_size = "quad_100",
                      ntrees = gmsum$nobs,
                      richness = gmsum$ngrps$sp[1])
  sdtab <- sdtab[rep(seq_len(nrow(sdtab)), each = 8), ]
  rownames(sdtab) <- NULL
  
  res <- cbind(sdtab, result)
kable(res, digits = 2)
# saving table ------------
save(res, file=here("workflow_example", "bci_models_outputs","time_models",
          "grow", "bci-quad_100-1-table.Rdata"))
```


## 2. Mortality time model

Loading the cleaned data.
```{r}
df <- readRDS(here("workflow_example", "bci_cleandata", "mortality.rds"))
data <- bind_rows(df, .id = "time")
data$quadrat <- data[,colnames(data) == "quad_100"]
```

Including column for the time interval between tree measurements, in years.
```{r}
data$y.interval <- as.numeric(data$interval/365) # interval years
```

Taking the subsample 1.
```{r}
data <- data[data$quad_100 %in% samples$bci$bci_1, ]
```

Models parameters.
```{r}
delta=0.95
ncores = 3 # number of chains
iter = 3000 # number of iterations per chain
warmup = 1000 # number of burning iterations discarded per chain
thin = 5 # thinning interval
```

Model.
```{r, eval=F}
gm <- brm(dead ~ 1 + (1|quadrat) + (1|sp) + (1|time) +
              (1|quadrat:sp) + (1|quadrat:time) + (1|sp:time) +
              offset(log(y.interval)), 
            family = bernoulli(link="cloglog"),
            data=data, 
            chains = ncores,
            control = list(adapt_delta = delta),
            cores = ncores,
            iter = iter, warmup=warmup,
            thin = thin,
            seed=T,
            save_pars = save_pars(group=FALSE),
            inits="0")
save(gm, file = here("workflow_example", "bci_models_outputs","time_models",
          "mort", "bci-quad_100-1-model.Rdata"))
```

The model is already saved in the `bci_models_output` folder.
```{r}
load(here("workflow_example", "bci_models_outputs","time_models",
          "mort", "bci-quad_100-1-model.Rdata"))
```

Summary table with posterior medians.
```{r}
gmsum <- summary(gm, robust=T) # median posterior
gmsum
```

Traceplots for the posterior distribution of parameters and chains mixture evaluation. Also saved as pdf A4 files.
```{r, eval=F}
pdf(file=here("workflow_example", "bci_models_outputs","time_models",
          "mort", "bci-quad_100-1-traceplot.pdf"), 
    paper="a4", height = 20, width=10,
    onefile = T)
plot(gm, ask=F, N=4)
dev.off()
```

```{r}
plot(gm, ask=F, N=4)
```

Table of model results.
```{r}
fix <- as.data.frame(gmsum$fixed)
  res <- as.data.frame(gmsum$fixed)
  res[1,] <- c(sqrt((pi^2)/6), rep(NA,6))
  
  quadrat <- as.data.frame(gmsum$random$quadrat)
  sp <- as.data.frame(gmsum$random$sp)
  time <- as.data.frame(gmsum$random$time)
  quadrat.sp <- as.data.frame(gmsum$random$`quadrat:sp`)
  quadrat.time <- as.data.frame(gmsum$random$`quadrat:time`)
  sp.time <- as.data.frame(gmsum$random$`sp:time`)
  
  result <- bind_rows(list(intercept = fix, 
                           quadrat = quadrat,
                           sp = sp,
                           time = time,
                           `quadrat:sp` = quadrat.sp,
                           `quadrat:time` = quadrat.time,
                           `sp:time` = sp.time,
                           Residual = res), .id="term")
  rownames(result) <- NULL
  
  
  result$levels = c(NA, gmsum$ngrps$quadrat[1],
                    gmsum$ngrps$sp[1],
                    gmsum$ngrps$time[1],
                    gmsum$ngrps$`quadrat:sp`[1],
                    gmsum$ngrps$`quadrat:time`[1],
                    gmsum$ngrps$`sp:time`[1],
                    gmsum$nobs)
  result$variance <- result$Estimate^2
  
  sdtab <- data.frame(sub = "1",
                      data = "all", 
                      fplot = "bci",
                      q_size = "quad_100",
                      ntrees = gmsum$nobs,
                      richness = gmsum$ngrps$sp[1])
  sdtab <- sdtab[rep(seq_len(nrow(sdtab)), each = 8), ]
  rownames(sdtab) <- NULL

  res <- cbind(sdtab, result)
kable(res, digits = 2)
# saving table ------------
save(res, file=here("workflow_example", "bci_models_outputs","time_models",
          "mort", "bci-quad_100-1-table.Rdata"))
```



## 3. Recruitment time model

Loading the cleaned data.
```{r}
df <- readRDS(here("workflow_example", "bci_cleandata", "recruitment.rds"))
data <- bind_rows(df, .id = "time")
data$quadrat <- data[,colnames(data) == "quad_100"]
```

Including column for the quadrat mean time interval in years.
```{r}
inter <- data %>% group_by(time, quadrat) %>% 
    summarise(y.interval = as.numeric(mean(interval, na.rm = T)/365))
# if interval is 0 or NaN, use mean of plot
inter$y.interval[is.nan(inter$y.interval) | inter$y.interval == 0] <- 
    mean(inter$y.interval, na.rm = T)
```

Taking the subsample 1
```{r}
data <- data[data$quad_100 %in% samples$bci$bci_1, ]
```

Models parameters.
```{r}
delta=0.95
ncores = 3 # number of chains
iter = 3000 # number of iterations per chain
warmup = 1000 # number of burning iterations discarded per chain
thin = 5 # thinning interval
```

Model.
```{r, eval=F}
gm <- brm(rec ~ 1 + (1|quadrat) + (1|sp) + (1|time) +
              (1|quadrat:sp) + (1|quadrat:time) + (1|sp:time) +
              offset(log(y.interval)), 
            family = bernoulli(link="cloglog"),
            data=data, 
            chains = ncores,
            control = list(adapt_delta = delta),
            cores = ncores,
            iter = iter, warmup=warmup,
            thin = thin,
            seed=T,
            save_pars = save_pars(group=FALSE),
            inits="0")
save(gm, file = here("workflow_example", "bci_models_outputs","time_models",
          "rec", "bci-quad_100-1-model.Rdata"))
```

The model is already saved in the `bci_models_output` folder.
```{r}
load(here("workflow_example", "bci_models_outputs","time_models",
          "rec", "bci-quad_100-1-model.Rdata"))
```

Summary table with posterior medians.
```{r}
gmsum <- summary(gm, robust=T) # median posterior
gmsum
```

Traceplots for the posterior distribution of parameters and chains mixture evaluation. Also saved as pdf A4 files.
```{r, eval=F}
pdf(file=here("workflow_example", "bci_models_outputs","time_models",
          "rec", "bci-quad_100-1-traceplot.pdf"), 
    paper="a4", height = 20, width=10,
    onefile = T)
plot(gm, ask=F, N=4)
dev.off()
```

```{r}
plot(gm, ask=F, N=4)
```

Table of model results.
```{r}
fix <- as.data.frame(gmsum$fixed)
  res <- as.data.frame(gmsum$fixed)
  res[1,] <- c(sqrt((pi^2)/6), rep(NA,6))
  
  quadrat <- as.data.frame(gmsum$random$quadrat)
  sp <- as.data.frame(gmsum$random$sp)
  time <- as.data.frame(gmsum$random$time)
  quadrat.sp <- as.data.frame(gmsum$random$`quadrat:sp`)
  quadrat.time <- as.data.frame(gmsum$random$`quadrat:time`)
  sp.time <- as.data.frame(gmsum$random$`sp:time`)
  
  result <- bind_rows(list(intercept = fix, 
                           quadrat = quadrat,
                           sp = sp,
                           time = time,
                           `quadrat:sp` = quadrat.sp,
                           `quadrat:time` = quadrat.time,
                           `sp:time` = sp.time,
                           Residual = res), .id="term")
  rownames(result) <- NULL
  
  
  result$levels = c(NA, gmsum$ngrps$quadrat[1],
                    gmsum$ngrps$sp[1],
                    gmsum$ngrps$time[1],
                    gmsum$ngrps$`quadrat:sp`[1],
                    gmsum$ngrps$`quadrat:time`[1],
                    gmsum$ngrps$`sp:time`[1],
                    gmsum$nobs)
  result$variance <- result$Estimate^2
  
  sdtab <- data.frame(sub = "1",
                      data = "all", 
                      fplot = "bci",
                      q_size = "quad_100",
                      ntrees = gmsum$nobs,
                      richness = gmsum$ngrps$sp[1])
  sdtab <- sdtab[rep(seq_len(nrow(sdtab)), each = 8), ]
  rownames(sdtab) <- NULL

  res <- cbind(sdtab, result)
kable(res, digits = 2)
# saving table ------------
save(res, file=here("workflow_example", "bci_models_outputs","time_models",
          "rec", "bci-quad_100-1-table.Rdata"))
```

